---
title: "Gene expression FCs normalized for stromal TIL"
author: "Maarten Slagter"
output:
  html_document:
    toc: true
    number_sections: true
    theme: united
---


```{r, warning=FALSE, echo=FALSE, cache = F}
setwd('~/TONIC')
source('.Rprofile')
options(mc.cores = 2)
knitr::opts_chunk$set(message = FALSE, cache = T, cache.lazy = F,
                      cache.comments = F, autodep = T, warning = FALSE,
                      results = 'hide', fig.keep = 'high',
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
library(tidybayes)
# devtools::install_github('paul-buerkner/brms')
library(brms)

# knitr::clean_cache()
# knitr::opts_chunk$get("cache.path")
# knitr::knit_global()
# setwd('~/Projects/TONIC')
source('R/init.R')
```

# Introduction

Due to relatively small patient numbers per treatment arm, spontaneous biases in
baseline factors could confound differences in the FCs of the genesets we've
decided to follow over time - those associated with clinical response at the on
nivolumab timepoint. Here I show that among the patients for whom NanoString
assays are available both On nivo and at Baseline, inter-arm differences in
stromal TIL% between the treatment arms are not detectable. Stromal TIL
additionally only seems weakly associated with the FCs of the median geneset
scores. As an effect of that, trying to correct for stromal TIL by explicitly
taking it along in a regression model does not alter the relative weight that is
assigned to the various treatment arms - as one would expect based on the
observed data.

# Methods

The log2 FC in the median of the 12-geneset values per patient was modelled as a
function of either the treatment arm alone (the baseline model) or both the
treatment arm and the baseline level of stromal TIL percentage. The latter could
be interpreted as a measure of the baseline immunological presence in the
assayed sample and hence as a proxy for the local immune system's capability of
inducing an inflamed state.

In a Bayesian framework, FCs were assumed to be Gaussian distributed with a mean
that is composed of the sum of arm and stromal TIL effects. Arm coefficients
were regularized by drawing them from a common $t$-distribution. For the prior
distribution on the stromal TIL, we also chose a $t_2$-distribution. TODO: check
and further specify exact parameters.

# No apparent relationship between baseline factors and expression FCs

```{r, warning=FALSE, echo=FALSE}
# source('R/test_associations.R')
source('R/bayes.R')
# debugonce(prepare_test_gene_set_difference)
dtf <- prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction')
dtf <- prep_comp_data(tp1 = 'Baseline', tp2 = 'On nivo')
```

```{r, warning=FALSE, echo=FALSE}
explore_confounder('s_til')
explore_confounder('ca15_3')
```

# Regression modelling of the observed FCs

```{r, warning=FALSE, echo=FALSE}
source('R/bayes.R')

## Test stability of inferences
dtf <- prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction')
dtf[, .N, patient]
dtf[, .N, arm]
# hlm_test <- run_sim(df_sigma_FC = 1, df_sigma_arm = 1,
#                data = dtf, adapt_delta = .99)
sr(hlm_test)

fn <- 'test.pdf'
pdf(fn)
pairs(hlm_test, las = 1)
dev.off()
sys_file_open(fn)
```

```{r, warning=FALSE, echo=FALSE}
source('R/bayes.R')

dtf <- prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction')

extract_rel_parms <- function(df_sigma_FC = 1, df_sigma_arm = 3) {
  hlm <- run_sim(df_sigma_FC = df_sigma_FC, df_sigma_arm = df_sigma_arm,
                 data = dtf, adapt_delta = .99)
  get_posterior_mean(hlm,
                     # grep('df_sigma_FC|df_sigma_arm|FC|normalized', names(hlm), value = T)) %>%
                     grep('normalized', names(hlm), value = T)) %>%
    # debug_pipe
    { .[, 'mean-all chains'] }
}

if (T) {
  hyper_param_scan <- expand.grid(df_sigma_FC = 1:6, df_sigma_arm = 1:6) %>%
    as.data.table
  hyper_param_scan %<>%
    { .[, extract_rel_parms(df_sigma_FC = df_sigma_FC,
                            df_sigma_arm = df_sigma_arm),
        by = .(df_sigma_FC, df_sigma_arm)] }
  hyper_param_scan$param_name <-
    names(extract_rel_parms(df_sigma_FC = 1, df_sigma_arm = 1))
  hyper_param_scan <- dcast(hyper_param_scan,
                            df_sigma_FC + df_sigma_arm ~ param_name,
                            value.var = 'V1')
  saveRDS(hyper_param_scan, file.path(rds_dir, 'hyper_param_scan.rds'))
} else {
  sr(hyper_param_scan)
}

ggplot(hyper_param_scan, aes(x = as.factor(df_sigma_FC),
                             y = as.factor(df_sigma_arm),
                             fill = `sd_arm[4]`)) +
 geom_tile() +
 scale_x_discrete(expand = c(0, 0)) +
 scale_y_discrete(expand = c(0, 0)) +
 theme(legend.text = element_text(size = 5), legend.direction = 'vertical',
       legend.position = 'right')
```

## Acount for baseline levels

```{r, warning=FALSE, echo=FALSE}
source('R/bayes.R')
hlm <- run_sim(data = data_pi,
  use_baseline = F, use_clinical_response = F)
LOO <- loo(hlm, cores = 46)
```

```{r, warning=FALSE, echo=FALSE}
# source('R/init.R')
source('R/bayes.R')
getDoParWorkers()
options(mc.cores = 10)

pi_BM_sims_t <- gen_all_model_fits(
  dtf = prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction'),
  iter = 5e5,
  stan_model = 'rmd/stan_model_t_version.stan')
saveRDS(pi_BM_sims_t, file.path(rds_dir, 'pi_BM_sims_t.rds'))

pi_BM_sims_n <- gen_all_model_fits(
  dtf = prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction'),
  iter = 5e5,
  stan_model = 'rmd/stan_model_n_version.stan')
saveRDS(pi_BM_sims_n, file.path(rds_dir, 'pi_BM_sims_n.rds'))
# pryr::size(pi_BM_sims_n)
```

```{r, warning=FALSE, echo=FALSE}
# source('R/init.R')
source('R/bayes.R')

pi_BM_sims_t_s <- gen_all_model_fits(
  dtf = prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction'),
  iter = 1e5,
  stan_model = 'rmd/stan_model_t_version.stan')
saveRDS(pi_BM_sims_t_s, file.path(rds_dir, 'pi_BM_sims_t_s.rds'))
```

```{r, warning=FALSE, echo=FALSE}
source('R/bayes.R')
pi_BM_sims_n_s <- gen_all_model_fits(
  dtf = prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction'),
  iter = 1e5,
  stan_model = 'rmd/stan_model_n_version.stan')
saveRDS(pi_BM_sims_n_s, file.path(rds_dir, 'pi_BM_sims_n_s.rds'))
```

```{r, warning=FALSE, echo=FALSE}
## 2019-01-31 15:49 Studying the variance of estimates rather informally
## 2019-02-01 15:24 I have increased the amount of iterations by 50x to reduce
## this variance
rbindlist(purrr::imap(pi_BM_sims, function(obj, n) {
  i_pars <- intersect(fitted_coefs(obj),
    c('FC_arm_normalized[1]', 'FC_arm_normalized[2]',
      'FC_arm_normalized[3]', 'FC_arm_normalized[4]'))
  if (length(i_pars) == 0) return(NULL)
  rstan::extract(obj, pars = i_pars) %>%
    as.data.table %>%
    { .[, lapply(.SD, mean)] } %>%
    cbind(., 'name' = n)
}))
## Run 1
#    FC_arm_normalized[1] FC_arm_normalized[2] FC_arm_normalized[3]
# 1:               -0.294               0.1227                0.881
# 2:               -0.335              -0.0888                0.700
# 3:               -0.244               0.1609                0.980
# 4:               -0.320              -0.0832                0.690
# 5:               -0.341              -0.0929                0.695
# 6:               -0.277              -0.0280                0.754
# 7:               -0.305              -0.0560                0.740
#    FC_arm_normalized[4]        name
# 1:                0.467         hlm
# 2:                0.429       hlm_b
# 3:                0.569       hlm_r
# 4:                0.472     hlm_b_r
# 5:                0.467   hlm_b_r_d
# 6:                0.335   hlm_b_r_l
# 7:                0.358 hlm_b_r_d_l
## Run 2
#    FC_arm_normalized[1] FC_arm_normalized[2] FC_arm_normalized[3]
# 1:               -0.286               0.1411                0.910
# 2:               -0.345              -0.0814                0.686
# 3:               -0.256               0.1683                0.979
# 4:               -0.333              -0.0749                0.718
# 5:               -0.329              -0.0730                0.710
# 6:               -0.264              -0.0188                0.756
# 7:               -0.313              -0.0601                0.726
#    FC_arm_normalized[4]        name
# 1:                0.477         hlm
# 2:                0.437       hlm_b
# 3:                0.613       hlm_r
# 4:                0.500     hlm_b_r
# 5:                0.501   hlm_b_r_d
# 6:                0.346   hlm_b_r_l
# 7:                0.381 hlm_b_r_d_l

## 2019-02-01 15:34 Two fits with student t distro and with
coefs <- rbindlist(c(
  study_covar_values(pi_BM_sims_t, name = 'student_t_1'),
  study_covar_values(pi_BM_sims_n, name = 'student_t_2')
), fill = T)
#     FC_arm_normalized.V1 FC_arm_normalized.V2 FC_arm_normalized.V3
#  1:               -0.296               0.1382                0.899
#  2:               -0.327              -0.0874                0.686
#  3:               -0.256               0.1629                0.968
#  4:               -0.310              -0.0745                0.712
#  5:               -0.326              -0.0907                0.686
#  6:               -0.267              -0.0235                0.761
#  7:               -0.286              -0.0371                0.734
#  8:               -0.293               0.1378                0.904
#  9:               -0.327              -0.0881                0.685
# 10:               -0.258               0.1618                0.964
# 11:               -0.314              -0.0777                0.708
# 12:               -0.333              -0.0866                0.682
# 13:               -0.267              -0.0243                0.759
# 14:               -0.290              -0.0375                0.732
#     FC_arm_normalized.V4   modelname  bc[1]   rc[1]  tc[1] lc[1]
#  1:                0.461 student_t_1     NA      NA     NA    NA
#  2:                0.442 student_t_1 -0.310      NA     NA    NA
#  3:                0.577 student_t_1     NA -0.2548     NA    NA
#  4:                0.484 student_t_1 -0.302 -0.0933     NA    NA
#  5:                0.499 student_t_1 -0.292 -0.0469 -0.133    NA
#  6:                0.360 student_t_1 -0.290 -0.0785     NA 0.373
#  7:                0.376 student_t_1 -0.273 -0.0309 -0.145 0.374
#  8:                0.464 student_t_2     NA      NA     NA    NA
#  9:                0.439 student_t_2 -0.310      NA     NA    NA
# 10:                0.576 student_t_2     NA -0.2577     NA    NA
# 11:                0.480 student_t_2 -0.303 -0.0910     NA    NA
# 12:                0.495 student_t_2 -0.288 -0.0465 -0.140    NA
# 13:                0.363 student_t_2 -0.290 -0.0751     NA 0.369
# 14:                0.374 student_t_2 -0.273 -0.0300 -0.147 0.377

rbindlist(study_covar_values(pi_BM_sims_n_s, name = 'normal'), fill = T)
#    FC_arm_normalized.V1 FC_arm_normalized.V2 FC_arm_normalized.V3
# 1:               -0.331                0.157                1.027
# 2:               -0.393               -0.247                0.645
# 3:               -0.285                0.198                1.133
# 4:               -0.375               -0.225                0.687
# 5:               -0.389               -0.233                0.667
# 6:               -0.345               -0.179                0.731
# 7:               -0.358               -0.186                0.709
#    FC_arm_normalized.V4 modelname  bc[1]   rc[1]   tc[1] lc[1]
# 1:                0.529    normal     NA      NA      NA    NA
# 2:                0.513    normal -0.490      NA      NA    NA
# 3:                0.717    normal     NA -0.3770      NA    NA
# 4:                0.575    normal -0.480 -0.1232      NA    NA
# 5:                0.583    normal -0.469 -0.0903 -0.0843    NA
# 6:                0.476    normal -0.459 -0.1078      NA 0.246
# 7:                0.482    normal -0.446 -0.0715 -0.0917 0.253
```

```{r, warning=FALSE, echo=FALSE}
if (!exists('pi_BM_sims_t'))
  pi_BM_sims_t <- readRDS(file.path(rds_dir, 'pi_BM_sims_t.rds'))

if (!exists('pi_BM_sims_n'))
  pi_BM_sims_n <- readRDS(file.path(rds_dir, 'pi_BM_sims_n.rds'))

coefs <- rbindlist(c(
  study_covar_values(pi_BM_sims_t, name = 'student_t'),
  study_covar_values(pi_BM_sims_n, name = 'normal')
), fill = T)
print(coefs)

```

```{r, warning=FALSE, echo=FALSE}
## Try to print posterior means only
# lapply(pi_BM_sims, function)
LOO_compares <- loo::compare(x = map(pi_BM_sims_t, loo))
extract_log_lik(pi_BM_sims[[1]])
```

```{r, warning=FALSE, echo=FALSE}
source('R/bayes.R')

if (F && !exists('pi_BM_sims_t'))
  pi_BM_sims_t <- readRDS(file.path(rds_dir, 'pi_BM_sims_t.rds'))

if (!exists('pi_BM_sims_n'))
  pi_BM_sims_n <- readRDS(file.path(rds_dir, 'pi_BM_sims_n.rds'))

fn <- file.path(img_dir, 'panel_bayesian_post_induction.pdf')

if (F) {
  plot_panel_layout(imap(pi_BM_sims_t, function(d, n)
      plot_params(p_dat = d, mn = n, save_plot = F)),
    filename = fn, ncol = 2, nrow = 3, labels = tolower(LETTERS),
    w = 19.5, h = 25, panel_padding = .5)
}
```

# Generate supplemental Figure 8

```{r, warning=FALSE, echo=FALSE}
data_pi <- prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction')
# data_pi <- prep_comp_data(tp1 = 'Baseline', tp2 = 'On nivo')
# data_pi[, lymphnode_only_disease :=
#   ifelse(lymphnode_only_disease,
#     'Lymph node only mets', 'Mets to other sites')]
# data_pi[, lymphnode_only_disease := factor(lymphnode_only_disease,
#   levels = c('Lymph node only mets', 'Mets to other sites'))]

# data_pi[, .N, by = .(arm, lines_of_therapy_for_metastatic_disease)]
# data_pi[, mean(lines_of_therapy_for_metastatic_disease), by = .(arm)]
# data_pi[, .('frac_LN_only' =
#   mean(lymphnode_only_disease == 'Mets to other sites')), by = .(arm)]

# CR_boxplot <- ggplot(data_pi,
#   aes(x = clinical_response, y = FC, fill = clinical_response)) +
#   geom_boxplot(outlier.size = 0, width = .5) +
#   ggbeeswarm::geom_beeswarm(dodge.width = 10, alpha = .6) +
#   facet_wrap(~arm, nrow = 5) +
#   coord_flip() +
#   scale_fill_manual(values = tonic_color_palettes[['clinical_response']]) +
#   scale_x_discrete(name = 'Clinical response') +
#   scale_y_continuous(name = 'log2FC inflammation score')
# p <- ggplot(data_pi, aes(x = clinical_response, y = FC_comp)) + geom_boxplot()
# p + facet_grid(~arm)
# data_pi[, .N, .(arm, clinical_response)][order(arm, clinical_response)]

ggplot2::theme_set(theme_ms(base_size = 8,
  panel.border = ggplot2::element_rect(colour = 'grey20', fill = NA, size = 1,
                                       linetype = 'solid')))

BLGE_scatter <- data_pi %>%
  ggplot(aes(x = Baseline, y = FC)) +
  # geom_point() + geom_smooth(method = 'lm',
  #   formula = y ~ x + 0) +
  geom_point() + geom_smooth(method = 'lm',
    formula = y ~ x, colour = 'red') +
  ylab('log2FC inflammation score') +
  xlab('Baseline inflammation score')

CR_boxplot <- ggplot(data_pi,
  aes(x = clinical_response, y = FC)) +
  geom_boxplot(outlier.size = 0, width = .5) +
  ggbeeswarm::geom_beeswarm(dodge.width = 1, alpha = .6, cex = 3) +
  geom_point(
    data = data_pi[, .('FC' = mean(FC)), by = .(clinical_response)],
    colour = 'red', alpha = .5, size = 2) +
  scale_x_discrete(name = 'Clinical response') +
  scale_y_continuous(name = 'log2FC inflammation score')

LN_boxplot <- ggplot(data_pi, aes(x = lymphnode_only_disease, y = FC)) +
  geom_boxplot(outlier.size = 0, width = .5) +
  ggbeeswarm::geom_beeswarm(dodge.width = 1, alpha = .6, cex = 3) +
  geom_point(data =
    data_pi[, .('FC' = mean(FC)), by = .(lymphnode_only_disease)],
    colour = 'red', alpha = .5, size = 2) +
  scale_x_discrete(name = 'Metastases restricted to lymph nodes') +
  scale_y_continuous(name = 'log2FC inflammation score')

TL_boxplot <- ggplot(data_pi,
  aes(x = lines_of_therapy_for_metastatic_disease > 0, y = FC)) +
  geom_boxplot(outlier.size = 0, width = .5) +
  ggbeeswarm::geom_beeswarm(dodge.width = 1, alpha = .6, cex = 4) +
  geom_point(data =
    data_pi[, .('FC' = mean(FC)),
       by = .(lines_of_therapy_for_metastatic_disease > 0)],
    colour = 'red', alpha = .5, size = 2) +
  scale_x_discrete(name = 'Palliative treatment prior to TONIC') +
  scale_y_continuous(name = 'log2FC inflammation score')
# print(TL_boxplot)
# plots <- list(BLGE_scatter, CR_boxplot, LN_boxplot, TL_boxplot)
# fn <- file.path(img_dir, 'panel_bayesian_covariates.pdf')
# plot_panel_layout(plots, filename = fn, ncol = 2, labels = tolower(LETTERS),
#                   w = 19.5, h = 20, panel_padding = .5)
```

```{r, warning=FALSE, echo=FALSE}
# devtools::load_all('~/libs/maartenutils')
# source('R/bayes.R')
library(gridExtra)
library(glue)

object_name <- 'pi_BM_sims_t_s'
object_name <- 'pi_BM_sims_n_s'
obj_ext <- gsub('.*_sims_(.*)$', '\\1', object_name)

if (!exists(object_name))
  assign(object_name,
    readRDS(file.path(rds_dir, sprintf('%s.rds', object_name))))
# with(get(object_name),
#   plot_params(p_dat = hlm_b_r_d_l, mn = 'post_induction_b_r_d_l'))
# rbindlist(study_covar_values(get(object_name), name = obj_ext), fill = T)
# get(object_name)[[1]]

tab <- rbindlist(study_covar_values(get(object_name), name = obj_ext), fill = T)
invisible(tab[, modelname := NULL])
```

```{r, warning=FALSE, eval=F, echo=FALSE}
as.list(expression(paste('Histogram of ' , hat(mu)), 
    'Bootstrap samples, Allianz'))
params <- sapply(tolower(treatment_arms[2:5]), function(x)
  atop(glue('\'Relative fold induction {x}\''), 
    glue('(mu[{x}]-mu[no~induction])')))
params <- sapply(tolower(treatment_arms[2:5]), function(x)
  atop(glue('\'Relative fold induction {x}\''), 
    glue('(mu[{x}]-mu[no~induction])')))
setnames(tab, c(params, 'b', 'c', 't', 'l'))
tab <- tab[, lapply(.SD, function(x) {
  x[is.na(x)] <- ''
  return(x)
})]
coef_table <- gridExtra::tableGrob(tab, 
  rows = NULL,
  theme = ttheme_default(base_size = 6, 
    colhead=list(fg_params = list(parse=TRUE))))
# plot(coef_table)
```

```{r, warning=FALSE, echo=FALSE}
param_explanation_table <- as.data.frame(matrix(c(
  'mu', 'Global mean fold change',
  'mu[arm]', 'Arm-level mean fold change',
  'mu[FC]', 'Expected individual fold change',
  'FC', 'Observed fold change',
  'sigma[arm]', '\'Standard deviation in\'~mu[arm]',
  'sigma[FC]', 'Standard deviation in FC',
  'b', 'Effect of baseline inflammation score',
  'S[BL]', 'Inflammation score at baseline',
  'c', '\'Effect\' of clinical response',
  'R', 'Indicator variable of clinical response',
  'l', 'Effect of prior palliative treatment',
  'L', 'Indicator variable of prior palliative treatment',
  'n', 'Effect of having metastases restricted to lymph nodes',
  'N', 'Indicator variable of lymph-node restricted metastases'
), byrow = T, ncol = 2))
colnames(param_explanation_table) <- c('Parameter', 'Description')

param_explanation_table <- gridExtra::tableGrob(
  param_explanation_table, 
  rows = NULL,
  theme = ttheme_default(base_size = 6, 
    core = list(fg_params = list(parse=TRUE)),
    padding = grid::unit(c(2, 2), 'mm'),
    colhead = list(fg_params = list(parse=TRUE))))
# rownames(param_explanation_table) <- NULL
# plot(param_explanation_table)
```

```{r, warning=FALSE, echo=FALSE}
source('R/bayes.R')

plots <-
  with(get(object_name), list(
    NA,
    param_explanation_table,
    plot_params(p_dat = hlm),
    BLGE_scatter,
    CR_boxplot, TL_boxplot, LN_boxplot,
    plot_params(p_dat = hlm_b_r_d_l, ncol = 2),
    # coef_table
    NA
  ))

fn <- file.path(img_dir, sprintf('panel_bayesian_revisions_%s.pdf', obj_ext))
plot_panel_layout(plots, filename = fn, ncol = 2,
  labels = tolower(LETTERS) %>% append('', after = 1),
  # labels = tolower(LETTERS),
  layout_mat = matrix(
    c(1, 1, 3,
      2, 2, 4,
      5, 6, 7,
      8, 8, 8,
      9, 9, 9), ncol = 3, byrow = T),
  w = 19.5, h = 27.7, panel_padding = .5, widths = c(.275, .275, .45))
```


```{r, warning=FALSE, echo=FALSE}
devtools::load_all(file.path('~/libs', 'maartenutils'))

plots <-
  list(prep_comp_data(tp1 = 'Baseline', tp2 = 'Post-induction') %>%
         ggplot(aes(x = Baseline, y = FC)) + geom_point() + geom_smooth() +
         ylab('log2 FC baseline vs. post-induction') +
         xlab('Baseline gene set score'),
       prep_comp_data(tp1 = 'Baseline', tp2 = 'On nivo') %>%
         ggplot(aes(x = Baseline, y = FC)) + geom_point() + geom_smooth() +
         ylab('log2 FC baseline vs. on nivolumab') +
         xlab('Baseline gene set score'),
       plot_params(hlm_pi, mn = 'post_induction', save_plot = F) +
         ggtitle('Post-induction vs. baseline'),
       plot_params(hlm_b_pi, mn = 'post_induction_use_baseline', save_plot = F) +
         ggtitle('Post-induction vs. baseline\n\\w baseline score'),
       plot_params(hlm, mn = 'on_nivo', save_plot = F) +
         ggtitle('On nivolumab vs. baseline'),
       plot_params(hlm_b, mn = 'on_nivo_use_baseline', save_plot = F) +
         ggtitle('On nivolumab vs. baseline\n\\w baseline score')
       )
# { align_plots(plotlist = ., 'hv') }

plot_panel_layout(plots, filename = file.path(img_dir, 'panel_bayesian.pdf'),
                  ncol = 2)
```

# Earlier mucking with the brms package

```{r, warning=FALSE, echo=FALSE}
brm_fc_s <- do.call('brm',
                  c(list(formula = bf(FC ~ 0 + (1 | arm), nl = F),
                          prior = set_prior('student_t(10, 0, 5)',
                                            class = 'sd', group = 'arm'),
                          data = dtf, family = gaussian()),
                    brms_settings[c('warmup', 'iter', 'chains', 'control')]))
stancode(brm_fcs[[3]])
standata(brm_fc_s)
```


```{r, warning=FALSE, echo=FALSE}
priors <- c("normal(0,10)"
stancode(brm_fc)
p_names <- names(brm_fc$fit@sim$samples[[1]])
```

```{r, warning=FALSE, echo=FALSE}
p <- stanplot(sigma_brm_fcs[[3]], pars = '^r_.*(arm)',
              type = 'intervals') + theme_ms()
p <- p + scale_y_discrete(limits = rev(grep('r_arm',
                                        names(brm_fc$fit@sim$samples[[1]]),
                                        value = T)),
                          labels = rev(treatment_arms))
p <- p + xlab('Credible interval for log2(FC)')
print(p)
ggsave(p, filename = file.path(img_dir, 'bayesian_fc.pdf'))
```



```{r, warning=FALSE, echo=FALSE}
posterior_samples(brm_fc, "b")
get_prior(FC ~ 0 + (0 + intercept | arm), data = dtf, family = gaussian())
get_prior(FC ~ 0 + intercept, data = dtf, family = gaussian())
```

```{r, warning=FALSE, echo=FALSE}
pp_check(brm_fc)
plot(brm_fc)
```

```{r, eval = F, brm2, warning=FALSE, echo=FALSE}
# source('R/bayes.R')
# devtools::load_all(file.path('~/libs', 'maartenutils'))
# debugonce(posterior_histogram)
brm_fc_stil <- do.call('brm',
                       c(list(formula = bf(FC ~ s_til + (1 | arm), nl = F),
                              data = dtf, family = gaussian()),
                              brms_settings[c('warmup', 'iter',
                                              'chains', 'control')]))
print(brm_fc_stil)
```

Comparing the leave-one-out-cross-validation performance, the simpler model
taking along just treatment arm actually prevails. This is not unexpected given
the weak relationship between stromal TIL and treatment arm in the assayed
samples. Stromal TIL is given a very weak estimate in the expanded model; the
models are virtually identical.

```{r, warning=FALSE, echo=FALSE}
loo(brm_fc, brm_fc_stil)
```

Assessing the posterior distributions of arm-coefficients in the models, we see
that the effects ascribed to Cisplatin are higher than that of Doxorubicin and
that this difference is slightly increased when accounting for Stromal TIL -
indeed stromal TIL levels are slightly higher in Dox arm, making the
contribution of the stromal TIL coefficient to the observed FCs slightly larger
here. Assuming validity of the data generating model and convergence of the
MCMC-sampling (highly probably given the convergence estimates and MCMC-traces),
we can estimate that the probability that Cisplatin induces higher FCs than
Doxorubicin is ~80%. I haven't done other comparisons yet, should I compare all
induction arms to the No induction arm?

```{r, warning=FALSE, echo=FALSE, fig.width = 14/2.54, fig.height = 10/2.54}
source('R/bayes.R')
# brm_fc
p_names <- names(brm_fc$fit@sim$samples[[1]])
# debugonce(posterior_histogram)

posterior_histogram(brms_object = brm_fc,
                    name = 'Model: Induction therapy',
                    binwidth = .05,
                    cols = tonic_color_palettes[['arm']],
                    x_lab = 'Contribution to FC',
                    # variables = c('Doxorubicin', 'Cisplatin', 'No.induction')) +
                    variables = gsub(' ', '\\.', treatment_arms)) +
  theme(legend.position = 'right') +
  ggtitle('Baseline model')
```

```{r, warning=FALSE, echo=FALSE}
p <- stanplot(brm_fc, pars = '^r_.*(arm)', type = 'intervals') + theme_ms()
p <- p + scale_y_discrete(limits = rev(grep('r_arm',
                                        names(brm_fc$fit@sim$samples[[1]]),
                                        value = T)),
                          labels = rev(treatment_arms))
p <- p + xlab('Arm FC relative to all patients (log2)')
print(p)
fn <- file.path(img_dir, 'bayesian_FC_intercepts.pdf')
ggsave(plot = p, file = fn,
       width = 8, height = 8, units = 'cm')
sys_file_open(fn)
```

```{r, warning=FALSE, echo=FALSE}
dtf <- extract_params(brm_fc,
               params = grep('r_arm', names(brm_fc$fit@sim$samples[[1]]),
                             value = T))
lapply(dtf, function(x) quantile(x, probs = c(0.025, .975)))
lapply(dtf, function(x) quantile(x, probs = c(0.05, .95)))
```

```{r, warning=FALSE, echo=FALSE, fig.width = 14/2.54, fig.height = 10/2.54}
posterior_histogram(brms_object = brm_fc_stil,
                    name = 'Model: Induction therapy + stromal TIL',
                    cols = tonic_color_palettes[['arm']],
                    x_lab = 'Contribution to FC',
                    variables = c('Doxorubicin', 'Cisplatin', 'No.induction')) +
  theme(legend.position = 'right') +
  ggtitle('Expanded model')
```



```{r, warning=FALSE, echo=FALSE}
compute_bayesian_p_val <- function(brms_object = brm_fc,
                                   name = '',
                                   offset_val = 1,
                                   group_name = 'induction_therapy',
                                   param_capture = '.*\\[(.*),Intercept\\]',
                                   x_lab = 'Contribution to response probability',
                                   cols = NULL,
                                   variables = c('Doxorubicin', 'Cisplatin')) {
  ## Subselect group variables
  p_names <- names(brms_object$fit@sim$samples[[1]])
  p_names <- grep(param_capture, p_names, value = T)
  p_names <- setNames(p_names,
                      tolower(gsub(param_capture, '\\1', p_names)))
  if (all(sapply(variables, is.null)))
    variables <- names(p_names)

  p_dat <- as.data.frame(extract_params(brms_object, p_names))
  comp_vals <- setdiff(1:length(p_names), offset_val)
  setNames(sapply(comp_vals, function(comp_val)
           mean(p_dat[, offset_val] < p_dat[, comp_val])),
           names(p_names)[comp_vals])
  sapply(setNames(comp_vals, names(p_names)[comp_vals]),
         function(x) quantile(p_dat[, x], probs = c(.025, .5, .975)))
}
compute_bayesian_p_val()
```
