---
title: "Correlates of T cell infiltration"
author: "Maarten Slagter"
output:
  html_document:
    toc: true
    number_sections: true
    theme: united
---

Determine which genes correlate well with T cell infiltration as determined by
Adaptive.  Then use this gene signature to determine an expression score and see
what correlates to this expression signature.

```{r, warning=FALSE, echo=FALSE, cache = F}
knitr::opts_chunk$set(message = FALSE, cache = T, cache.lazy = F,
                      cache.comments = F, autodep = T, warning = FALSE,
                      results = 'hide', fig.keep = 'high',
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# knitr::clean_cache()
# knitr::opts_chunk$get("cache.path")
# knitr::knit_global()
setwd('~/Projects/TONIC')
source('R/init.R')
source('R/load_rna_dat.R')
```

```{r, warning=FALSE, echo=FALSE}
patient_labels_tmp <-
  patient_labels[timepoint %in% timepoints[c(1,2)], 
                 .(patient, timepoint, adaptive_t_cells)] %>%
  unique %>%
  { .[naturalorder(patient), ] }

# rna_sample_annotation[patient_labels_tmp[, .(patient, timepoint)], .N]
setkey(rna_sample_annotation, patient, timepoint)
patient_labels_tmp$rna_cf_number <- 
  rna_sample_annotation[patient_labels_tmp[, .(patient, timepoint)], cf_number]

patient_labels_tmp %<>% { .[!apply(., 1, function(x) any(is.na(x)))] }
```

```{r, warning=FALSE, echo=FALSE}
entrez_table <- readRDS(file.path(p_root, 'rds', 'entrez_table.rds'))
setkey(entrez_table, hgnc_symbol)

t_cell_corrs <- 
  ## Order samples identically as in patient_labels_tmp
  rna_read_counts_salmon_tmm_M[, patient_labels_tmp$rna_cf_number, with = F] %>%
  { apply(., 1, function(x) cor(x, patient_labels_tmp$adaptive_t_cells, 
                                method = 'spearman')) } %>%
  setNames(rownames(rna_read_counts_salmon_tmm_M))

covs <- 
  rna_read_counts_salmon_tmm_M %>%
  { apply(., 1, function(x) sd(x, na.rm = T) / mean(x, na.rm = T)) } %>%
  setNames(rownames(rna_read_counts_salmon_tmm_M))

# plot(ecdf(abs(t_cell_corrs)))

## Select genes associated with Adaptive T cell infiltration that vary over the
## samples
t_cell_genes <- names(t_cell_corrs[abs(t_cell_corrs) >= .3]) %>%
  setdiff(NA) %>%
  intersect(names(covs[covs >= 1]))

t_dat <- data.table(hgnc_symbol = names(t_cell_corrs), 
                    corr = t_cell_corrs,
                    cova = covs,
                    selected = names(t_cell_corrs) %in% t_cell_genes) %>%
  mutate(cova_bin = cut(cova, 
                        breaks = quantile(cova, 
                                          probs = seq(0, 1, by = .25)))) %>%
  controlled_merge(entrez_table[, .(hgnc_symbol, description)]) %>%
  { .[!is.na(corr)] } %>%
  arrange(corr) %>%
  mutate(cor_rank = rank(corr, ties.method = 'max')) %>%
  { mutate(., cor_rank = ifelse(corr > 0, 
                                cor_rank - .[corr <= 0, .N], 
                                .[corr <= 0, .N] - cor_rank - 1)) } 

t_dat %>%
  maartenutils::write_tsv(output_file = file.path('ext', 
                                                  't_cell_correlated_genes.tsv'))
```

```{r, warning=FALSE, echo=FALSE}
ggplot(t_dat[!is.na(cova_bin)], aes(x = corr, fill = cova_bin)) + 
  geom_histogram(bins = 200) +
  scale_fill_manual(name = 'Coefficient of variation',
                    values = rev(gen_color_vector('Spectral', 4))) +
  theme(legend.position = c(.95, .95), legend.direction = 'vertical',
        legend.justification = c(1, 1)) +
  scale_x_continuous(name = 'correlation with Adaptive T cell infiltration',
                     expand = c(0, 0)) +
  scale_y_continuous(name = '# genes', expand = c(0, 0))
```
  

```{r, warning=FALSE, echo=FALSE}
## Test for associations with T cell infiltration
search_term <- 'WNT'
search_term <- 'DKK\\d'
search_term <- '^CD8[A|B]$'
search_term <- '^CD4$'
search_term <- '^FOXP3$'
t_dat %>% filter(grepl(search_term, hgnc_symbol))
t_dat %>% filter(grepl(search_term, hgnc_symbol)) %>% select(corr) %>% sum
```

It seems as if both CD8 and CD4 (FoxP3) subsets increase with T cell
infiltration, I should test whether the correlation is indeed that clear.

```{r, warning=FALSE, echo=FALSE}
patient_labels_tmp %<>%
  maartenutils::controlled_merge(
    compute_gene_set_score(timepoint = timepoints[c(1,2)], 
                           gene_symbols = t_cell_genes,
                           sum_func = sum) %>% 
    dplyr::rename(rna_t_cell = gs_score))
```

If continueing along this track, perhaps first compute optimal weighting of each
gene for T cell infiltration.

What are the most variant genes?

Annotate GO functions

```{r, warning=FALSE, echo=FALSE}
## Downloaded from http://www.geneontology.org/page/download-go-annotations
go_terms <- fread(file.path(data_dir, 'goa_human.gaf'), header = F)
setnames(go_terms, c('uni', 'bs', 'hgnc_symbol', 'empty', 'go_term', 'go_ref',
                     'go_type', 'ref_link', 'some_letter', 'gene_description',
                     'more_nonsense', 'protein', 'taxon', 'date', 'source',
                     'rd1', 'rd2'))
go_terms[, .(hgnc_symbol, go_term, go_ref)]
```

```{r, warning=FALSE, echo=FALSE}
pacman::p_load(httr)
url  <- "http://api.epdb.eu"
path <- "eurlex/directory_code"
raw.result <- GET(url = url, path = path)
```

```{r, warning=FALSE, echo=FALSE}
pacman::p_load(httr)
url  <- "www.ebi.ac.uk/QuickGO/api/"
path <- "annotation/search"
raw.result <- GET(url = url, path = path)
```

```{r, warning=FALSE, echo=FALSE}
## Annotate annotation to GO IDs
select(GO.db, keys = 'GO:0000001', columns = c('GOID', 'DEFINITION',
                                               'ONTOLOGY', 'TERM'))

```

```{r, warning=FALSE, echo=FALSE}
search_term = '^RPL'

explore_in_browser(arrange(t_dat, -cova))

t_dat %>% 
  arrange(-cova) %>% 
  mutate(order_rank = seq(.N, 1)) %>%
  filter(grepl(search_term, hgnc_symbol)) %$%
  { c('rank_sum' = sum(order_rank), 
      'expected_rank_sum' = length(order_rank) * 21190 / 2,
      'normalized_rank_sum' = sum(order_rank) / (length(order_rank) * 21190 / 2)) } 
```

```{r, warning=FALSE, echo=FALSE}
graphics.off()
source('R/plotting_rna.R')
plot_scatter_cor(x_var = 'adaptive_t_cells',
                 y_var = 'rna_t_cell',
                 dtf = patient_labels_tmp,
                 cor_method = 'spearman',
                 point_alpha = .5,
                 axis_labeller = tonic_cap,
                 outlier_label_var = 'cf_number',
                 position = 'topleft') 
```

```{r, warning=FALSE, echo=FALSE}
# colnames(rna_t_dat)
# rownames(rna_t_dat)
plot_scatter_cor(x_var = 'FOXP3',
                 y_var = 'CD8A',
                 dtf = transpose_data.table(rna_read_counts_salmon_tmm_M) %>%
                         rownames_to_column('cf_number') %>%
                         dplyr::select_('cf_number', 'FOXP3', 'CD8A'),
                 cor_method = 'spearman',
                 point_alpha = .5,
                 axis_labeller = tonic_cap,
                 outlier_label_var = 'cf_number',
                 position = 'topleft') 
```

```{r, warning=FALSE, echo=FALSE}
plot_scatter_cor(x_var = 'adaptive_t_cells',
                 y_var = 'rna_t_cell',
                 dtf = patient_labels_tmp,
                 cor_method = 'spearman',
                 point_alpha = .5,
                 axis_labeller = tonic_cap,
                 outlier_label_var = 'cf_number',
                 position = 'topleft') 
```

```{r, warning=FALSE, echo=FALSE}
hist(patient_labels_tmp[, adaptive_t_cells], breaks = 100)
```

Horrendous reproduction of T cell infiltration relative order. Should I just
order samples based on adaptive T cells and only cluster genes?

```{r, warning=FALSE, echo=FALSE}
devtools::load_all('~/libs/GSEAgenesets')
source('R/rna.R')

plot_gene_set(gs_name = 'T_cell_correlated',
              # gene_symbols = t_cell_genes,
              gene_symbols = t_dat[abs(corr) > .5, hgnc_symbol],
              # setdiff(names(t_cell_corrs[abs(t_cell_corrs) >= .7]), NA),
              cf_numbers = arrange(patient_labels_tmp, adaptive_t_cells)$rna_cf_number,
              cf_ordered = T,
              exp_mat = rna_read_counts_salmon_tmm_M,
              sample_normalisation = 'lib_size',
              # sample_normalisation = NULL,
              gene_normalisation = F,
              hclustfun = 'complete',
              distfun = 'euclidean',
              col_vars = c('timepoint', 'arm',
                           'clinical_response', 'adaptive_t_cells'),
              timepoints = c('Baseline', 'Post-induction'))
```
