---
title: "Identifying clinical responders from RNASeq data in the TONIC trial"
author: "Maarten Slagter"
output: html_document
---

```{r, cache = F, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = F, cache.lazy = F,
                      cache.comments = F, autodep = T, warning = FALSE,
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
```


```{r, warning=FALSE, echo=FALSE, include=F}
source('R/init.R')
source('R/load_rna_dat.R')
source('R/ML_RNASeq.R')
```

# Check quality of data

## Other approaches

```{r, warning=FALSE, echo=FALSE}
apply(tpms_salmon, 2, compute_evenness)
apply(tpms_salmon, 2, median)
```

# Glmnet approach

```{r, warning=FALSE, echo=FALSE}
glmnet_class_test_rnaseq(tp = 'Baseline', type = 'genes')
```

```{r, warning=FALSE, echo=FALSE}
glmnet_class_test_rnaseq(tp = 'Post-induction', type = 'genes')
```

```{r, warning=FALSE, echo=FALSE}
source('R/ML_RNASeq.R')
layout(matrix(1:2, ncol = 2, nrow = 1, byrow = T))
r_b <- glmnet_class_test_rnaseq(tp = 'Baseline', type = 'genes', alphas = 1)
r_pi <- glmnet_class_test_rnaseq(tp = 'Post-induction', type = 'genes', alphas = 1)
```

```{r, warning=FALSE, echo=FALSE}
plot(r_b$glmnet.fit)
r_b %>%
  { coef(.$glmnet.fit, .$lambda.min) } %>%
  { setNames(.[which(. != 0)], rownames(.)[which(. != 0)]) }

r_pi %>%
  { coef(.$glmnet.fit, .$lambda.min) } %>%
  { setNames(.[which(. != 0)], rownames(.)[which(. != 0)]) } %>%
  { .[order(abs(.))] }
```

# SVM approach to training

```{r, warning=FALSE, echo=FALSE}
source('R/ML_RNASeq.R')
t_dat <- prepare_ml_dat_rnaseq(tp = 'Post-induction', type = 'genes',
                               # exp_mat = rna_read_counts_salmon_tmm)
                               exp_mat = tpms_salmon)
s2n_scores <- as.matrix(t_dat$t_dat) %>%
  # { apply(., 2, function(x) ggsea::ggsea_s2n(x, as.matrix(t_dat$responses))) }
  { apply(., 2, function(x) ggsea::ggsea_s2n(as.matrix(x), as.matrix(t_dat$responses))) }
  # { apply(., 2, function(x) mean(x)) }
  # { apply(., 2, function(x) str(x)) }
s2n_scores %>% { .[abs(.) >= 1] }
```

```{r, warning=FALSE, echo=FALSE}
ctrl <- trainControl(method = 'repeatedcv',   # 10fold cross validation
                     repeats = 1,				      # do 5 repititions of cv
                     # Use AUC to pick the best model
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

str(t_dat$t_dat)
svm.tune <- train(x = t_dat$t_dat,
                  y = t_dat$responses,
                  method = 'gaussprPoly',
                  tuneLength = 9,					        # 9 values of the cost function
                  # preProc = c('center','scale'),  # Center and scale data
                  metric = 'ROC',
                  trControl = ctrl)
```

```{r, warning=FALSE, echo=FALSE}
ctrl <- trainControl(method = 'repeatedcv',   # 10fold cross validation
                     repeats = 1,				      # do 5 repititions of cv
                     # Use AUC to pick the best model
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

svm.tune <- train(x = t_dat$t_dat,
                  y = t_dat$responses,
                  method = 'svmRadial',           # Radial kernel
                  tuneLength = 9,					        # 9 values of the cost function
                  preProc = c('center','scale'),  # Center and scale data
                  metric = 'ROC',
                  trControl = ctrl)
```

```{r, warning=FALSE, echo=FALSE}
plot(svm.tune)
confusionMatrix(svm.tune)
```

# Naive/reference model performance

Naive model classification performance
Classify according to response rate

```{r, warning=FALSE, echo=FALSE}
p <- mean(as.logical(responses))
p;^2 + (1 - p)^2
2 * p * (1 - p)
curve(expr = x^2 + (1 - x)^2, from = 0, to = 1)
# curve(expr = 2 * x * (1 - x), from = 0, to = 1)
```

# PCA approach

Response seems to be partially explained by the first two principal components
on therapy samples.
For timepoints later than 1, we should probably be looking at fold changes
instead.

```{r, warning=FALSE, echo=FALSE}
ml_dat <- prepare_ml_dat(tp = 'On nivo')
plot_PCA_axes(tp = 'on_nivo', dat = ml_dat[[1]], color_var = 'response',
              labellings = ml_dat[[2]])

ml_dat <- prepare_ml_dat(tp = 'Post-induction')
plot_PCA_axes(tp = 'post_induction', dat = ml_dat[[1]], color_var = 'response',
              labellings = ml_dat[[2]])

ml_dat <- prepare_ml_dat(tp = 'Baseline')
plot_PCA_axes(tp = 'baseline', dat = ml_dat[[1]], color_var = 'response',
              labellings = ml_dat[[2]])

setkey(patient_labels, filename)
ml_dat <- prepare_ml_dat(tp = 'Post-induction')
plot_PCA_axes(tp = 'post_induction', dat = ml_dat[[1]], color_var = 'arm',
              labellings = patient_labels[rownames(ml_dat[[1]]), arm])

setkey(patient_labels, filename)
ml_dat <- prepare_ml_dat(tp = 'Baseline')
plot_PCA_axes(tp = 'baseline', dat = ml_dat[[1]], color_var = 'arm',
              labellings = patient_labels[rownames(ml_dat[[1]]), arm])
```

# Random forest approach

Given that gene sets associated with clinical response at timepoint 2 were also
already higher at baseline (when combining all treatment arms), it seems
plausible that responders can be identified at baseline based on these gene
sets.

```{r, cache = T, dependson, warning=FALSE, echo=FALSE}
source('R/ML.R')
rfs <- plyr::llply(timepoints, function(tp) {
  perform_rf(tp = tp)
}, .parallel = !local_run)
```

```{r, warning=FALSE, echo=FALSE}
plots <- lapply(rfs, plot_importance)
plot_panel_layout(plots, ncol = 3, filename = NULL,
                  w = 20, h = 20, label_size = 8)
```

Use clinical information to train predictor

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
library(maartenutils)

x_dat <- clinical_annotation[, .(induction_therapy, who_performance_status,
                                 s_til, ca15_3, response_bin)] %>%
  maartenutils::filter_na_rows()
invisible(x_dat[, response_bin := factor(response_bin)])

rf_default <- caret::train(x = as.data.frame(select(x_dat, -response_bin)),
  y = unlist(select(x_dat, response_bin)),
  method = "rf",
  tuneGrid = expand.grid(.mtry=ceiling(sqrt(ncol(x_dat - 1)))),
  trControl = trainControl(method="repeatedcv",
  number = nrow(x_dat),
  savePredictions = T,
  repeats=10))
```

Kappa of zero indicates no added benefit over a 'null' classifier which simply
classifies randomly based on incidence.

```{r, warning=FALSE, echo=FALSE}
str(rf_default)
selectedIndices <- rf_default$pred$mtry == 3
str(rf_default)
rf_default$pred
library(caret)
library(pROC)
head(rf_default$pred)

plot.roc(rf_default$pred$obs[selectedIndices],
         rf_default$pred$M)

```

# Information content in gene set scores

Collinearity much less of a problem here

```{r, warning=FALSE, echo=FALSE}
sr(geneset_scores_tmm <- gen_gene_set_score_matrix(sum_func = median,
                           exp_mat = rna_read_counts_salmon_tmm,
                           sets = filter_gmt('h.all', 'HALLMARK'),
                           log_transform = 'log2'))
```

```{r, warning=FALSE, echo=FALSE}
cf_numbers <- rna_sample_annotation[timepoint == 'Post-induction', cf_number]

gs_mat <- dcast(geneset_scores_tmm[cf_number %in% cf_numbers],
      formula = patient ~ gene_set,
      value.var = 'value')
row_names <- rna_sample_annotation[match(cf_numbers, cf_number), patient]
row_names <- gs_mat[, patient]
rownames(gs_mat) <- row_names
gs_mat[, patient := NULL]
gs_mat[, cf_number := NULL]
gs_mat <- as.matrix(gs_mat)
rownames(gs_mat) <- row_names

resp_vec <- patient_labels[match(rownames(gs_mat), patient), clinical_response]
missing_idx <- which(is.na(resp_vec))
resp_vec <- factor(resp_vec[-missing_idx], levels = c('R', 'NR'))
gs_mat <- gs_mat[-missing_idx, ]
cf_numbers <- cf_numbers[-missing_idx]
```

```{r, warning=FALSE, echo=FALSE}
## Raw expression mat
re_mat <- t(rna_read_counts_salmon_tmm_M)
re_mat <- re_mat[match(rownames(re_mat), cf_numbers), ]
re_mat <- re_mat[!is.na(rownames(re_mat)), ]
colnames(re_mat) <- rownames(rna_read_counts_salmon_tmm_M)
rownames(re_mat) <- rna_sample_annotation[match(rownames(re_mat), cf_number), patient]
```

```{r, warning=FALSE, echo=FALSE}
library(XML)
doc.html = htmlTreeParse('https://rdrr.io/cran/caret/man/models.html',
                         useInternal = TRUE)
doc.value <- xpathApply(doc.html, '//h2|//p', xmlValue)
doc.html.value <- xpathApply(doc.html, '//h2|//p')
```

```{r, warning=FALSE, echo=FALSE}
train_caret <- function(weight_responders = 4,
                        train_methods = c('AdaBag', 'elm', 'naive_bayes'),
                        exp_mat = gs_mat) {
  ctrl <- trainControl(method = 'repeatedcv',   # 10fold cross validation
                       repeats = 10,				      # do 5 repititions of cv
                       # Use AUC to pick the best model
                       summaryFunction = twoClassSummary,
                       classProbs = TRUE)
  res <- lapply(train_methods, function(l_method) {
    message(l_method)
    # rownames(re_mat)
    # colnames(re_mat)
    model.tune <- train(x = exp_mat,
                        y = resp_vec,
                        weights = (weight_responders - 1) *
                          as.numeric(resp_vec == 'R') + 1,
                        method = l_method,    # Radial kernel
                        tuneLength = 9,				# 9 values of the cost function
                        preProc = c(),
                        metric = 'ROC',
                        trControl = ctrl)
    return(model.tune)
  })
  return(reds)
}
models <- train_caret(exp_mat = gs_mat)
```

```{r, warning=FALSE, echo=FALSE}
res <- caret::confusionMatrix(svm.tune, mode = 'prec_recall')
```

# Dimension reduction

## NMF decomposition of expression matrix

```{r, warning=FALSE, echo=FALSE}
# source('~/antigenic_space/bin/install_packages.R')
devtools::load_all('~/libs/GSEAgenesets')
# pacman::p_load('fastICA')
setwd('~/Projects/TONIC')
source('R/init.R')
source('R/load_rna_dat.R')
source('R/rna.R')
```

```{r, warning=FALSE, echo=FALSE}
pacman::p_load(NMF)
sr(aout <- nmf(rna_read_counts_salmon_tmm_M_ensg, 40))
w <- aout@fit@W
h <- aout@fit@H
approxa <- w %*% h
## Element-wise comparison
approx_quality <- 
  { (approxa - rna_read_counts_salmon_tmm_M_ensg) /
  rna_read_counts_salmon_tmm_M_ensg } %>% 
  as.matrix
approx_quality %<>% { which(!eps(rna_read_counts_salmon_tmm_M_ensg, 0)) }
ind <- which(eps(rna_read_counts_salmon_tmm_M_ensg, 0), arr.ind = T)
approx_quality[ind[1, 1], ind[1, 2]]
approx_quality[ind] <- NA
rownames(approx_quality) <- rownames(rna_read_counts_salmon_tmm_M_ensg)
gene_qual <- apply(approx_quality, 1, 
      function(x) quantile(x, probs = c(.1, .5, .9), na.rm = T))
sample_qual <- apply(approx_quality, 2, 
      function(x) quantile(x, probs = c(.1, .5, .9), na.rm = T))
inspect_mat(gene_qual)
inspect_mat(sample_qual)
```

```{r, warning=FALSE, echo=FALSE}
rownames(w) <- rownames(rna_read_counts_salmon_tmm_M_ensg)
w[1, ]
hist(log10(w[, 6] + 1), breaks = 200)
hist(apply(w, 2, sum), breaks = 20)
w[, 1] %>% { .[order(-.)] }
# index <- which(!eps(rna_read_counts_salmon_tmm_M_ensg, 0), arr.ind = T)
# approx_quality[]
```

## PCA

```{r, warning=FALSE, echo=FALSE}
# rownames(rna_read_counts_salmon)
# dimnames(rna_read_counts_salmon)[[1]] <- rownames(rna_read_counts_salmon)
source('R/rna.R')
pca <- run_pca(dtf = rna_read_counts_salmon_tmm_M_ensg)
pca <- run_pca(dtf = rna_read_counts_salmon_tmm_M_ensg, 
               filter_covariates = F)

hist(log10(pca$scale), breaks = 400)
apply(pca$x, 2, squared_sum) %>% { ./sum(.) } %>%
  named_vec_to_dt(name = 'PC', value = 'exp_var_prop') %>%
  # dplyr::mutate(PC = factor(PC, levels = unique(PC))) %>%
  dplyr::mutate(PC = as.numeric(gsub('PC', '', PC))) %>%
  dplyr::mutate(cum_var = cumsum(exp_var_prop)) %>%
  { qplot(x = PC, y = cum_var, data = .) } +
  theme(panel.grid.major = element_line(size = .5, color = 'grey90'))
hist(pca$rotation, breaks = 300)
```

```{r, warning=FALSE, echo=FALSE}
pca$rotation[, 2] %>%
  { .[order(abs(.), decreasing = T)] } %>%
  { .[abs(.) >= .02] }
```

```{r, warning=FALSE, echo=FALSE}
library(ggbiplot)

setkey(patient_labels, patient, timepoint)

color_var <-
patient_labels[rna_sample_annotation[match(rownames(pca$x), cf_number),
                                     .(patient, timepoint)], adaptive_t_cells]

g <- ggbiplot(pca, choices = 1:2, obs.scale = 1, var.scale = 1,
              groups = color_var,
              var.axes = F,
              arrow = F,
              ellipse = F, circle = F)
g <- g + scale_color_continuous(name = 'Adaptive T-cells')
g <- g + theme(legend.direction = 'horizontal',
               legend.key.size = unit(5, 'mm'),
               legend.text = element_text(size = 6),
               legend.position = 'top')
print(g)
```

```{r, warning=FALSE, echo=FALSE}
devtools::load_all(file.path('~/libs', 'maartenutils'))
source('R/GSEA_plotting.R')

for (idx in 1:10) {
  rna_sample_annotation %<>%
    controlled_merge(named_vec_to_dt(pca$x[, idx],
                                     name_var = 'cf_number',
                                     value_var = sprintf('PC%d_loading', idx)),
                     by_cols = 'cf_number',
                     dup_priority = 'a')

  if (F) {
    plot_pairwise_relationships(
      y_var = sprintf('PC%d_loading', idx),
      # y_var_trans = function(x) { log10(x + 1),
      dtf = rna_sample_annotation,
      var_labeller = tonic_cap,
      blacklist_vars = c('cf_number', 't_number', 'one_of_best_samples', 'fn',
                         grep('PC\\d+_loading', colnames(rna_sample_annotation),
                              value = T)),
      filename = file.path(img_dir, sprintf('PC%d_correlates.pdf', idx)))
  }
}
# hist(rna_sample_annotation$PC1_loading, breaks = 200)
```

Liver samples seem enriched in PC1

```{r, warning=FALSE, echo=FALSE}
rna_sample_annotation[, .(PC1_loading, cf_number,
                          baseline_biopsy_site), by = patient] %>%
  arrange(-PC1_loading)
```

PAM50 and PC2 seem correlated as well

```{r, warning=FALSE, echo=FALSE}
summary(pca)
rna_sample_annotation[, .(PC1_loading, cf_number,
                          pam50_subtype), by = patient] %>%
  arrange(-PC1_loading)
```

```{r, warning=FALSE, echo=FALSE}
qplot(data = rna_sample_annotation, x = pam50_subtype, y = adaptive_t_cells,
      geom = 'boxplot') + 
  ggpubr::stat_compare_means()
```

### PCs associated with PC2 loadings

Seemed like an interesting way to detect T cell exclusion mechanisms.
Unfortunately no PC loadings are strongly associated with PC2.

```{r, warning=FALSE, echo=FALSE}
library(corrplot)
rot_mat <- pca$rotation
rownames(rot_mat) <- gsub('\\.\\d+$', '', rownames(rot_mat))

sr(GO_terms_db)
GO_terms_db <- GO_terms_db[evidence != 'IEA']
setkey(GO_terms_db, ensembl_gene_id)

## PC2 is negatively associated with T cell levels
## Any factor positively associated with it will indirectly decrease T cell
## levels
cor(rot_mat, use = 'everything', method = 'pearson')[, 2]
cor(rot_mat, use = 'everything', method = 'spearman')[, 2]
plot_scatter_cor('PC1', 'PC2', dtf = as.data.frame(rot_mat)) + geom_smooth()
hist(rot_mat[, 1], breaks = 200)
goi <- sort(rot_mat[, 1], decreasing = T) %>% { .[. >= 0.04] }
goi <- sort(rot_mat[, 2], decreasing = T) %>% { .[. >= 0.02] }
goi <- sort(rot_mat[, 5], decreasing = T) %>%
goi <- rot_mat[, 5] %>% 
  { .[squared_sum_norm(.) > quantile(squared_sum_norm(.), probs = .995)] }

mean(names(goi) %in% GO_terms_db[, ensembl_gene_id])
explore_in_browser(GO_terms_db[names(goi)])
explore_in_browser(GO_terms_db[names(goi)][, .N, by = term][order(-N)])
```

```{r, warning=FALSE, echo=FALSE}
## Show variation on variable scores on PCs
goi <- c('WNT7A', 'SERPINB9')
GO_terms_db[gene_symbol %in% goi] %>% 
  unique(by = 'gene_symbol') %$% 
  { set_rownames(rot_mat[rownames(rot_mat) %in% ensembl_gene_id, ], 
                 gene_symbol) } %>% 
  { apply(., 1, squared_sum_norm) } %>%
  { pheatmap(., cluster_rows = F, cluster_cols = F, fontsize_row = 6) }
```

What other variables are big on PC7?

```{r, warning=FALSE, echo=FALSE}
rot_mat[, 7] %>% 
  { .[squared_sum_norm(.) > quantile(squared_sum_norm(.), probs = .99)] } %>%
  named_vec_to_dt %>%
  set_colnames(c('ensembl_gene_id', 'PC7_loading')) %>%
  controlled_merge(GO_terms_db[, .(ensembl_gene_id, gene_symbol)]) %>%
  arrange(PC7_loading)
```

```{r, warning=FALSE, echo=FALSE}
recon <- pca$rotation %*% pca$x
summary(pca)
str(pca$rotation) 
hist(pca$rotation, breaks = 200) 
hist(pca$x, breaks = 200) 
str(pca$x)
squared_sum <- function(vec) sum(vec^2)
squared_sum_norm <- function(vec) vec^2 / squared_sum(vec)
apply(pca$x, 2, squared_sum) %>% { 100 * ./sum(.) }
hist(pca$x[, 2]^2 / sum(pca$x[, 2]^2), breaks = 40)

## Only plot samples most heavily scored on any given PC
pca$x[, 2] %>% { .[which(squared_sum_norm(.) > .02)] } %>%
  named_vec_to_dt(name = 'cf_number', value = 'PC2_score') %>%
  controlled_merge(rna_sample_annotation[, .(cf_number, adaptive_t_cells)]) %>%
  { plot_scatter_cor(x_var = 'PC2_score', y_var = 'adaptive_t_cells', 
                     dtf = .) }

  # { qplot(x = 'PC2_score', y = 'adaptive_t_cells', data = .) }
lookup_cf <- function(cf_numbers) {
  tolower(rna_sample_annotation$cf_number) %in% tolower(cf_numbers)
  tolower(patient_labels$cf_number) %in% tolower(cf_numbers)
}
cf_numbers <- pca$x[, 2][c(which.min(pca$x[, 2]), which.max(pca$x[, 2]))]
```

```{r, warning=FALSE, echo=FALSE}
library(pheatmap)
graphics.off()
# dev.off()
loadings_mat <- pca$x[, 2:20]
setkey(rna_sample_annotation, cf_number)
ann_dat <- rna_sample_annotation[rownames(loadings_mat), 
                                 .(cf_number, 
                                   clinical_response, 
                                   # timepoint,
                                   baseline_biopsy_site)] %>% as.data.frame
stopifnot(all(rownames(loadings_mat) == ann_dat$cf_number))
colnames(ann_dat) %<>% 
  mapvalues('baseline_biopsy_site', 'tissue') %>%
  mapvalues('response', 'clinical_response')
ann_dat$cf_number <- NULL

aheatmap(loadings_mat, annRow = ann_dat, 
         breaks = 0,
         annColors = tonic_color_palettes[setdiff(colnames(ann_dat), 'cf_number')])
         # color = gen_color_vector('Spectral', 5))
```

