---
title: "NanoString data analyses in the TONIC trial"
author: "Maarten Slagter"
output:
  html_document:
    toc: true
    number_sections: true
    theme: united
---

```{r, cache = F, warning=FALSE, echo=FALSE}
setwd('~/TONIC')
knitr::opts_chunk$set(message = FALSE, cache = F, cache.lazy = F,
                      cache.comments = F, autodep = T, warning = FALSE,
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
source('R/init.R')
```

# Parallel coordinates plots

## Geneset level

Which induction arms most strongly upregulate gene signatures associated with
response?

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
source('R/plotting_adaptive.R')
source('R/plotting_nanostring.R')
source('R/test_associations.R')
colour_var = NULL
colour_var = 'patient'
colour_var = 'arm'
colour_var = 'response'
colour_var = 'clinical_response'

# debugonce(plot_parallel_coords_geneset)
# debugonce(prep_geneset_parallel_coords)
# debugonce(plot_parallel_coords)


plots <- lapply(danaher_scores.m[, auto_name(unique(variable))], function(x) {
  p <- plot_parallel_coords_geneset(x,
                                    colour_var = colour_var,
                                    facet_var = 'arm')
  return(p)
})
# print(plots)
```

```{r, warning=FALSE, echo=FALSE}
filename <- file.path(img_dir, sprintf('parallel_%s.pdf',
                                       ifelse(is.null(colour_var), 'ungrouped',
                                              colour_var)))

p <- maartenutils::plot_panel_layout(plots, ncol = 1, nrow = 3,
                                     filename = filename,
                                     labels = NULL,
                                     w = 22, h = 25,
                                     label_size = 8)
# sys_file_open(filename)
```

```{r, eval=F, warning=FALSE, echo=FALSE}
source('R/plotting_nanostring.R')
colour_var = 'clinical_response'
x <- danaher_scores.m[, auto_name(unique(variable))][1]
plot_parallel_coords_geneset(x, colour_var = colour_var, facet_var = 'arm')
```


```{r, warning=FALSE, echo=FALSE}
source('R/plotting_adaptive.R')
source('R/plotting_nanostring.R')
source('R/test_associations.R')
# debugonce(plot_parallel_coords_geneset)
# debugonce(plot_parallel_coords)
for (sig in c('tis', 'ifn.gamma', 'cytotoxicity')) {
  plot_parallel_coords_geneset(sig, colour_var = 'clinical_response', 
                               median_line_alpha = .8,
                               point_alpha = .35,
                               facet_var = 'arm')
  ggsave(file = file.path(plot_dir, sprintf('%s_parallel_coords.png', sig)), 
         width = 19, height = 10, units = 'cm')
}
```


```{r, eval = F, warning=FALSE, echo=FALSE}
source('R/plotting_nanostring.R')

gen_parallel_coord_plots <-
  function(facet_var = 'arm', colour_var = 'clinical_response') {
  plots <- lapply(danaher_scores.m[, auto_name(levels(variable))],
                  function(gene_set) {
    sum_dat <- prep_geneset_parallel_coords(gene_set = gene_set,
                                            colour_var = colour_var,
                                            facet_var = facet_var)
    p_dat <- danaher_scores.m[variable == gene_set & !is.na(value)]
    p <- plot_parallel_coords(p_dat = p_dat, sum_dat = sum_dat,
                              # palette_name = 'FantasticFox',
                              facet_var = facet_var,
                              point_alpha = .6,
                              sum_colors = T,
                              colour_var = colour_var, 
                              title = gene_set) +
      theme(legend.position = 'none')
    return(p)
  })

  filename <- file.path(img_dir, sprintf('parallel_%s.pdf',
                                         ifelse(is.null(colour_var), 'ungrouped',
                                                colour_var)))
  p <- maartenutils::plot_panel_layout(plots, ncol = 1, nrow = 4,
                                       filename = filename,
                                       labels = NULL,
                                       w = 22, h = 25)

  # filename <- file.path(img_dir, sprintf('parallel_%s_highlighted.pdf',
  #                                        ifelse(is.null(colour_var), 'ungrouped',
  #                                               colour_var)))
  # p <- maartenutils::plot_panel_layout(
  #   list(plots[['tis']], plots[['glycolytic.activity']]),
  #   ncol = 1, nrow = 2,
  #   filename = filename,
  #   labels = NULL,
  #   w = 22, h = 25 / 2)
  return(plots)
}

# plots <- gen_parallel_coord_plots(facet_var = 'arm', colour_var = 'patient')
plots <- gen_parallel_coord_plots(facet_var = 'arm', colour_var = 'clinical_response')
```

```{r, eval = F, warning=FALSE, echo=FALSE}
# source('R/init.R')
# source('R/plotting_adaptive.R')
# source('R/plotting_nanostring.R')
# devtools::load_all(file.path('~/libs', 'maartenutils'))
library(grid)
plot_panel_layout(lapply(sig_gene_sets, function(gs) {
                           p <- plot_parallel_coords_geneset(gene_set = gs)
                           if (gs != sig_gene_sets[1])
                             p <- p + ggplot2::theme(legend.position = 'none')
                           else
                             p <- p + ggplot2::theme(legend.position = 'right')
                           return(p)
                         }),
                  filename = file.path('plots', 'parallel_sig_signatures.pdf'),
                  labels = NULL,
                  w = 17.4, h = 25, ncol = 3)
```

# Heatmap of all values for response associated sigs
 
Heatmap of single timepoint expression values

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_adaptive.R')
source('R/plotting_nanostring.R')
source('R/test_associations.R')
timepoint <- 'On nivo'
t_dat <- test_gene_set_association(
  gene_set = danaher_scores.m[, levels(variable)], timepoint = timepoint)
# t_dat <- t_dat[log2FC > 0 & p_val <= 0.05]
# t_dat <- t_dat[p_val <= 0.05]
sig_gene_sets <- setdiff(t_dat[p_val <= 0.05, as.character(variable)], '')
setkey(t_dat, variable)
# t_dat[gsub('_on.nivo', '', var_names)]
```

```{r, warning=FALSE, echo=FALSE}
source('R/GSEA_plotting.R') #renew tonic_cap()

gen_all_patient_nanostring_heatmap <- function(timepoint) {
  bfn <- file.path(plot_dir, sprintf('nanostring_heatmap_%s.pdf', 
                                     paste(timepoint, collapse = '-')))
  timelabel_danaher <- c('Baseline' = 'baseline', 
                         'On nivo' = 'on.nivo', 
                         'Post-induction' = 'post.induction')[timepoint]
  gene_sets <- sig_gene_sets
  gene_sets <- danaher_scores.m[, levels(variable)]

  var_names <- sapply(timelabel_danaher, function(tp) 
                      sprintf('%s_%s', gene_sets, tp))

  p_dat <- danaher_scores[, c(var_names, 'patient'), with = F] %>%
    { .[complete.cases(.)] } %>%
    mutate(patient = sprintf('pat_%s', patient)) %>%
    controlled_merge(patient_labels[, .(patient, arm, 
                                        clinical_response, pfs_binned)])

  if (length(timelabel_danaher) == 2) {
    for (i in 1:nrow(var_names)) {
      p_dat[, (gsub('_.*$', '', var_names[i, 2])) :=
            get(var_names[i, 2]) - get(var_names[i, 1])]
      p_dat[, (var_names[i, 1]) := NULL]
      p_dat[, (var_names[i, 2]) := NULL]
    }
    var_names %<>% { gsub('_.*$', '', .) } %>% as.vector %>% unique
  }

  NMF::aheatmap(t(p_dat[, var_names, with = F]),
                scale = 'row',
                border_color = 'grey80',
                labRow = var_names %>% 
                  as.vector %>%
                  { gsub('\\.', ' ', .) } %>%
                  { gsub('_on.nivo', '', .) } %>%
                  { gsub('_post.induction', '', .) } %>%
                  { gsub('_baseline', '', .) } %>%
                  { tonic_cap(., cap_first_word_only = F) },
                labCol = p_dat[, patient],
                # treeheight = 30,
                # cellheight = 50,
                # cellwidth = 50,
                annCol = list('Clinical response' = t(p_dat[,
                                                      clinical_response]),
                              'PFS' = t(p_dat[, pfs_binned]),
                              'Induction arm' = t(p_dat[, arm])),
                annColors = list('Clinical response' =
                                 tonic_color_palettes[['clinical_response']],
                                 'Induction arm' =
                                 tonic_color_palettes[['arm']],
                                 'PFS' = tonic_color_palettes[['pfs_binned']]),
                width = 20/2.54, height = 10/2.54, 
                # filename = file.path(plot_dir, 'heatmap_on_nivo.pdf'))
                filename = bfn)
                # filename = NA)
  invisible()
}

gen_all_patient_nanostring_heatmap(timepoint = 'On nivo')
gen_all_patient_nanostring_heatmap(timepoint = 'Post-induction')
gen_all_patient_nanostring_heatmap(timepoint = 'Baseline')
gen_all_patient_nanostring_heatmap(timepoint = c('Baseline', 'On nivo'))
gen_all_patient_nanostring_heatmap(timepoint = c('Post-induction', 'On nivo'))
gen_all_patient_nanostring_heatmap(timepoint = c('Baseline', 'Post-induction'))
```

# Heatmaps of median FCs with and without p-values

```{r, warning=FALSE, echo=FALSE, fig.width = 9, fig.height = 12}
source('R/plotting_adaptive.R')
source('R/plotting_nanostring.R')
source('R/test_associations.R')
sig_gene_sets <- setdiff(t_dat[p_val <= 0.05, as.character(variable)], '')
m <- rbindlist(lapply(auto_name(sig_gene_sets), function(gs) {
  test_gene_set_difference(gene_set = gs,
                           tp1 = 'Baseline',
                           tp = 'Post-induction')
}), fill = T)

invisible(m[p_val > .05, p_val := NA])
p <- plot_p_values_induction(m)
print(p)
ggsave(plot = p, filename = file.path(plot_dir,
                           'induction_fc_p_values_baseline_post_induction.pdf'),
       width = 10, height = 13, units = 'cm')
```

```{r, warning=FALSE, echo=FALSE, fig.width = 9, fig.height = 12}
devtools::load_all(file.path('~/libs', 'maartenutils'))
invisible(m[, p_val := NA])
pa <- plot_p_values_induction(filter(m, arm != 'All arms'))
print(pa)
ggsave(plot = pa, filename = file.path(plot_dir,
                           'induction_fc_p_values_baseline_post_induction.png'),
       width = 10, height = 13, units = 'cm')
```

```{r, warning=FALSE, echo=FALSE, fig.width = 9, fig.height = 12}
m <- rbindlist(lapply(auto_name(sig_gene_sets), function(gs) {
  test_gene_set_difference(gene_set = gs,
                           tp1 = 'Baseline',
                           tp = 'On nivo')
}), fill = T)

invisible(m[p_val > .05, p_val := NA])
p <- plot_p_values_induction(m)
print(p)
ggsave(plot = p, filename = file.path(plot_dir,
                            'induction_fc_p_values_baseline_on_nivo.pdf'),
       width = 10, height = 13, units = 'cm')
```

```{r, warning=FALSE, echo=FALSE, fig.width = 9, fig.height = 12}
invisible(m[, p_val := NA])
pa <- plot_p_values_induction(filter(m, arm != 'All arms'))
print(pa)
ggsave(plot = pa, filename = file.path(plot_dir,
                            'induction_fc_p_values_baseline_on_nivo.png'),
       width = 10, height = 13, units = 'cm')
# ggsave(plot = p, filename = file.path(plot_dir,
#                             'induction_fc_p_values_baseline_on_nivo.png'),
#        width = 10, height = 13, units = 'cm')
```

# NanoString correlations with clinical response

```{r, warning=FALSE, echo=FALSE}
timepoint = 'Baseline'
source('R/test_associations.R')
# lapply(danaher_signature_genes[, auto_name(unique(signature_name))],
#        function(sig) {
#          test_gene_set_association(gene_set = sig, timepoint = timepoint)
#        })
p_dat <- test_gene_set_association(
  gene_set = danaher_scores.m[, unique(variable)], timepoint = timepoint)
p_dat <- test_gene_set_association(
  gene_set = danaher_scores.m[, unique(variable)], timepoint = 'Post-induction')
p_dat <- test_gene_set_association(
  gene_set = danaher_scores.m[, unique(variable)], timepoint = 'On nivo')
# p_dat[p_val <= .05]
```

```{r, warning=FALSE, echo=FALSE}
plot_volcano <- function(gene_sets = danaher_scores.m[, unique(variable)],
                         timepoint = 'Baseline') {
  p_dat <- test_gene_set_association(gene_set = gene_sets,
                                     timepoint = timepoint)
  ggplot(p_dat, aes(x = log2FC, y = -log10(p_val), 
                    label = tonic_cap(variable))) +
    geom_hline(yintercept = -log10(.05),
               color = 'darkred', linetype = 'dashed') +
    geom_point() +
    ggrepel::geom_label_repel(data = p_dat[p_val <= .05 | 
                              variable %in% c('tis', 'ifn.gamma', 'pd1')],
                              min.segment.length = 0,
                              max.iter = 1e5,
                              force = 20) +
    geom_vline(xintercept = 0, color = 'grey70', linetype = 'dashed') +
    scale_y_continuous(name = 'Unadjusted p-value', trans = 'identity',
                       breaks = c(seq(0, 10, by = .5), -log10(0.05)),
                       limits = c(0, 3.5),
                       # labels = function(x) {
                       #   sapply(x, function(xi) {
                       #     if (xi == 0) {
                       #       return('0')
                       #     } else if (xi == -log10(0.05)) {
                       #       return('0.05')
                       #     } else {
                       #       return(parse(text = sprintf('10^{-%.2f}', xi)))
                       #     }
                       #   })}) +
                       labels = function(x) {
                             return(parse(text = sprintf('10^{-%.2f}', x)))
                         }) +
    # ggtitle(timepoint) +
    scale_x_continuous(name = sprintf('Fold difference\nresponders and non-responders (%s)',
                                      tolower(timepoint)),
                       limits = c(-2.7, 2.7))
}

plots <- list(
  plot_volcano(timepoint = 'Baseline'),
  plot_volcano(timepoint = 'Post-induction'),
  plot_volcano(timepoint = 'On nivo'))
cowplot::plot_grid(plots[[1]], plots[[2]], plots[[3]], ncol = 3)
ggsave(filename = file.path(img_dir,
                            'gene_sigs_response_volcanoes_all_timepoints.pdf'),
       width = 29, height = 11, units = 'cm')
cowplot::plot_grid(plots[[1]], plots[[3]], ncol = 2)
ggsave(filename = file.path(img_dir, 'gene_sigs_response_volcanoes.pdf'),
       width = 24, height = 11, units = 'cm')
```


# Correlations between RNA and NanoString

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
plots <- lapply(danaher_signature_genes[, unique(signature_name)],
                function(sig) {
  compare_nano_rna_sigs(sig = sig)
}) %>% { .[!sapply(., is.null)] }

maartenutils::plot_panel_layout(plots, ncol = 3, nrow = 4,
  labels = NULL,
  plot_direct = F,
  ref_panel_idx = 3 * 3,
  clear_redundant_labels = T,
  filename = file.path(plot_dir, sprintf('rna_nano_cors_%s.pdf', colour_var)),
  w = 17.4, h = 25, label_size = 8)
```


## The same but for individual genes

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')

rna_nano_cors <- rbindlist(lapply(auto_name(unique(test_genes)), function(gs) {
  t_dat <- correlate_nano_rna(genes = gs)
  if (null_dat(t_dat)) return(NULL)
  cors <- t_dat[, .('pearson' = cor(rna_value, nano_value,
                                    method = 'pearson'),
                    'spearman' = cor(rna_value, nano_value,
                                     method = 'spearman'),
                    rna_q50, nano_q50),
                by = c('gene_symbol', 'timepoint')]
  return(cors)
}), fill = T)

rna_nano_cors_sigs <- rbindlist(lapply(auto_name(danaher_scores.m[, levels(variable)]),
                                  function(sig) {
  t_dat <- correlate_nano_rna(sig = sig, genes = NULL)
  if (null_dat(t_dat)) return(NULL)
  cors <- t_dat[, .('pearson' = cor(rna_value, nano_value,
                                    method = 'pearson', use = 'complete'),
                    'spearman' = cor(rna_value, nano_value,
                                     method = 'spearman', use = 'complete'),
                    rna_q50, nano_q50, 'gene_symbol' = sig),
                by = c('timepoint')]
  return(cors)
}), fill = T)

comb_dat <- rbind(mutate(rna_nano_cors, 'type' = 'Single gene'),
                  mutate(rna_nano_cors_sigs, 'type' = 'Gene signature'))
```

```{r, warning=FALSE, echo=FALSE}
plot_dist <- function(dtf = comb_dat[timepoint == 'Baseline'],
                      cor_measure = 'spearman') {
  mycols <- c(maartenutils::darken('lightblue', factor = 1),
              maartenutils::darken('lightblue', factor = 2.5)) %>%
   setNames(dtf[, unique(type)])

  p1 <- ggplot(dtf,
               aes_string(x = cor_measure, y = '(..count..)/sum(..count..)',
                          fill = 'type')) +
    geom_histogram(binwidth = .01, color = 'black', position = 'stack') +
    scale_fill_manual(name = '', values = mycols) +
    xlab(sprintf('%s correlation coefficient', tonic_cap(cor_measure))) +
    ylab('Relative frequency') +
    theme(legend.position = 'top')

  p2 <- ggplot(dtf, aes_string(x = '1', y = cor_measure, fill = 'type')) +
    geom_boxplot(color = 'black') +
    scale_fill_manual(name = '', values = mycols) +
    coord_flip() +
    xlab(' ') +
    ylab(' ') +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          axis.line = element_blank(), legend.position = 'none')

  cowplot::plot_grid(p1, p2, align = 'v', rel_heights = c(7, 1), ncol = 1)
}

plot_dist(cor_measure = 'spearman')
```

```{r, warning=FALSE, echo=FALSE}
plot_dist(cor_measure = 'spearman')
ggsave(filename = file.path(plot_dir, 'spearman_nano_vs_rna_single_genes.pdf'))
plot_dist(cor_measure = 'pearson')
ggsave(filename = file.path(plot_dir, 'pearson_nano_vs_rna_single_genes.pdf'))
```

Plot relationship between correlation and average expression. Are lowly expressed
genes enriched amongst the genes with low concordance between the two
methodologies?


```{r, warning=FALSE, echo=FALSE}
ggplot(unique(comb_dat, by = c('gene_symbol')), 
       aes(x = nano_q50, y = spearman)) + geom_point()
```

```{r, warning=FALSE, echo=FALSE}
library(kableExtra)

unique(comb_dat, by = c('gene_symbol'))[abs(spearman) <= .1] %>%
  .[order(spearman)] %>%
  knitr::kable(format = 'html') %>%
  kableExtra::kable_styling()
```


