# Init

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
source('R/load_rna_dat.R')
```

# Compute FCs a priori

Abandoned for now

```{r, warning=FALSE, echo=FALSE}
source('R/load_rna_dat.R')
test_NA <- function(vec) is.na(vec) | vec == 'NA'
rna_sample_annotation_def <-
  copy(rna_sample_annotation[!test_NA(clinical_response)])
rna_sample_annotation_def[, timepoint := droplevels(timepoint)]
```

```{r, warning=FALSE, echo=FALSE}
cpms <- as.matrix(rna_read_counts)
rownames(cpms) <- ensgs
cpms <- cpms[!eps(apply(cpms, 1, var), 0, 1e-1), ]
## TODO include design matrix
cpms <- limma::voom(cpms)
```

```{r, warning=FALSE, echo=FALSE}
FCs <- rbindlist(lapply(unique(rna_sample_annotation$patient), function(patient) {
  cf_bl <- rna_sample_annotation[patient == parent.frame(3)$patient &
                        timepoint == 'Baseline', cf_number]
  cf_pi <- rna_sample_annotation[patient == parent.frame(3)$patient &
                        timepoint == 'Post-induction', cf_number]
  if (is.null(cf_bl) || is.null(cf_pi)) return(NULL)
  fcs <- as.list(cpms$E[, cf_pi] - cpms$E[, cf_bl], 'patient' = patient)
  return(fcs)
}), fill = T)
dim(FCs)
t(FCs)
FCs[1:5, 1:5]
```

# Paired testing

```{r, devel, eval = F, warning=FALSE, echo=FALSE}
source('R/GSEA_funcs_paired.R')
res <- gsea_wrapper(gene_sets = HALLMARK_pathways[1:3],
                    patients = rna_sample_annotation[arm == 'No induction',
                                                     unique(patient)],
                    resp_exp = 'timepoint',
                    nperm = 10,
                    abs = F,
                    fn_extra = '')
```

```{r, warning=FALSE, echo=FALSE}
## This one is generated on Coley
# devtools::load_all('~/libs/GSEAgenesets')
# devtools::load_all(file.path('~/libs', 'maartenutils'))
source('R/GSEA_plotting.R')
res <- readRDS(file = 'rds2/GSEA_paired.rds')
plot_es(res, legend.position = 'bottom', ptitle = 'paired',
        # filename = sprintf('plots2/%s/GSEA_%s.pdf',
        #                    format(Sys.time(), "%Y_%b_%d"), 'paired'),
        subset_var = NULL, x_var = 'arm')
```

## Heatmaps for leading edge genes.

Not very informative for the paired analysis

```{r, warning=FALSE, echo=FALSE}
# source('R/load_rna_dat.R')
library(GSEAgenesets)
# devtools::load_all('~/libs/GSEAgenesets')
source('R/rna.R')
source('R/GSEA_plotting.R')
# debugonce(preprocess_rna)
# debugonce(plot_gene_set)
plot_leading_edge_heatmap(res, gsea_name = 'paired',
                          timepoint = timepoints[1:2],
                          fdr_thresh = .25)
```

## Parallel coordinates of leading edge gene set scores

```{r, warning=FALSE, echo=FALSE}
res <- readRDS(file = 'rds2/GSEA_paired.rds')
dtf <- compute_leading_edge_gene_scores(res, fdr_thresh = .25)
plot_leading_edge_gene_scores(dtf, allow_nonsensical = T,
                              res_name = 'paired_FC_leading_edge')
```

# Unpaired testing

## Baseline

```{r, devel, eval = F, warning=FALSE, echo=FALSE}
source('R/GSEA_funcs_paired.R')
res <- gsea_wrapper(gene_sets = HALLMARK_pathways[1:3],
                    patients = rna_sample_annotation[arm == 'No induction',
                                                     unique(patient)],
                    resp_exp = 'timepoint',
                    nperm = 10,
                    abs = F,
                    fn_extra = '')
```

```{r, warning=FALSE, echo=FALSE}
res <- readRDS(file = 'rds2/GSEA_unpaired_baseline.rds')
plot_es(res, legend.position = 'bottom', ptitle = 'unpaired',
        subset_var = NULL, x_var = 'arm')
```


### Heatmaps for leading edge genes.

```{r, warning=FALSE, echo=FALSE}
plot_leading_edge_heatmap(res, gsea_name = 'unpaired',
                          timepoint = timepoints[1],
                          distfun = 'euclidean',
                          fdr_thresh = .25)
```

### Parallel coordinates for leading edge genes.

```{r, warning=FALSE, echo=FALSE}
source('R/rna.R')
res <- readRDS(file = 'rds2/GSEA_unpaired_baseline.rds')
dtf <- compute_leading_edge_gene_scores(res, fdr_thresh = .25)
plot_leading_edge_gene_scores(dtf, allow_nonsensical = T,
                              res_name = 'BL_r_vs_nr_leading_edge')
```


## Post-induction

```{r, devel, eval = F, warning=FALSE, echo=FALSE}
source('R/GSEA_funcs_paired.R')
res <- gsea_all_arms(gene_sets = HALLMARK_pathways,
                     patients = rna_sample_annotation[!is.na(clinical_response),
                                                      unique(patient)],
                     allowed_timepoints = 'Post-induction',
                     gene_score_fn = my_unpaired_WC_test,
                     resp_exp = 'clinical_response',
                     paired_test = F,
                     nperm = 1e1,
                     abs = F,
                     fn_extra = '')
```

```{r, warning=FALSE, echo=FALSE}
res <- readRDS(file = 'rds2/GSEA_unpaired_postinduction.rds')
plot_es(res, legend.position = 'bottom', ptitle = 'unpaired',
        subset_var = NULL, x_var = 'arm')
```

```{r, warning=FALSE, echo=FALSE}
plot_leading_edge_heatmap(res, gsea_name = 'unpaired',
                          timepoint = timepoints[2],
                          distfun = 'euclidean',
                          fdr_thresh = .25)
```

### Parallel coordinates of leading edge gene set scores

```{r, warning=FALSE, echo=FALSE}
# source('R/load_rna_dat.R')
library(GSEAgenesets)
source('R/rna.R')
dtf <- compute_leading_edge_gene_scores(res, fdr_thresh = .25)
```

```{r, warning=FALSE, echo=FALSE}
source('R/rna.R')
plot_leading_edge_gene_scores(dtf, allow_nonsensical = T,
                              res_name = 'PI_r_vs_nr_leading_edge')
```
