# Init

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
source('R/load_rna_dat.R')
```

# Compute FCs a priori

Abandoned for now

```{r, warning=FALSE, echo=FALSE}
source('R/load_rna_dat.R')
test_NA <- function(vec) is.na(vec) | vec == 'NA'
rna_sample_annotation_def <-
  copy(rna_sample_annotation[!test_NA(clinical_response)])
rna_sample_annotation_def[, timepoint := droplevels(timepoint)]
```

```{r, warning=FALSE, echo=FALSE}
cpms <- as.matrix(rna_read_counts)
rownames(cpms) <- ensgs
cpms <- cpms[!eps(apply(cpms, 1, var), 0, 1e-1), ]
## TODO include design matrix
cpms <- limma::voom(cpms)
```

```{r, warning=FALSE, echo=FALSE}
FCs <- rbindlist(lapply(unique(rna_sample_annotation$patient), function(patient) {
  l_patient <- patient
  cf_bl <- rna_sample_annotation[patient == l_patient &
                                 timepoint == 'Baseline', cf_number]
  cf_pi <- rna_sample_annotation[patient == l_patient &
                                 timepoint == 'Post-induction', cf_number]
  if (is.null(cf_bl) || is.null(cf_pi)) return(NULL)
  fcs <- as.list(cpms$E[, cf_pi] - cpms$E[, cf_bl], 'patient' = l_patient)
  return(fcs)
}), fill = T)
dim(FCs)
t(FCs)
FCs[1:5, 1:5]
```

# Paired testing

```{r, devel, eval = F, warning=FALSE, echo=FALSE}
source('R/GSEA_funcs_paired.R')
res <- gsea_wrapper(gene_sets = HALLMARK_pathways,
                    patients = rna_sample_annotation[arm == 'No induction',
                                                     unique(patient)],
                    exp_mat = rna_read_counts_salmon_tmm_M,
                    resp_exp = 'timepoint',
                    nperm = 100,
                    abs = F,
                    fn_extra = '')
```

```{r, warning=FALSE, echo=FALSE}
## This one is generated on Coley
# devtools::load_all('~/libs/GSEAgenesets')
# devtools::load_all(file.path('~/libs', 'maartenutils'))
source('R/GSEA_plotting.R')
# res <- readRDS(file = 'rds2/GSEA_paired.rds')
res <- readRDS(file = 'rds2/GSEA_paired_TMM.rds')

# filename <- sprintf('plots2/%s/GSEA_%s.pdf', format(Sys.time(), "%Y_%b_%d"), 'paired')
# dirname(filename)
dplyr::filter(res, arm != 'All arms' & grepl('HALLMARK', GeneSet)) %>%
  plot_es(legend.position = 'right', ptitle = 'paired_hallmark',
          subset_var = NULL, x_var = 'arm')

dplyr::filter(res, arm != 'All arms' & !grepl('HALLMARK', GeneSet)) %>%
  plot_es(legend.position = 'right', ptitle = 'paired_nonhallmark',
          subset_var = NULL, x_var = 'arm')
```

```{r, warning=FALSE, echo=FALSE}
res <- readRDS(file = 'rds2/GSEA_paired_TMM_senescence.rds')
dplyr::filter(res, arm != 'All arms') %>%
  plot_es(legend.position = 'right', ptitle = 'paired_senescence',
          # filename = sprintf('plots2/%s/GSEA_%s.pdf',
          #                    format(Sys.time(), "%Y_%b_%d"), 'paired'),
          subset_var = NULL, x_var = 'arm')
```

## Heatmaps for leading edge genes.

Not very informative for the paired analysis

```{r, warning=FALSE, echo=FALSE}
# source('R/load_rna_dat.R')
library(GSEAgenesets)
# devtools::load_all('~/libs/GSEAgenesets')
source('R/rna.R')
source('R/GSEA_plotting.R')
# debugonce(preprocess_rna)
# debugonce(plot_gene_set)
plot_leading_edge_heatmap(res, gsea_name = 'paired',
                          timepoint = timepoints[1:2],
                          fdr_thresh = .25)
```

## Parallel coordinates of leading edge gene set scores

```{r, warning=FALSE, echo=FALSE}
res <- readRDS(file = 'rds2/GSEA_paired.rds')
dtf <- compute_leading_edge_gene_scores(res, fdr_thresh = .25)
plot_leading_edge_gene_scores(dtf, allow_nonsensical = T,
                              res_name = 'paired_FC_leading_edge')
```

# Unpaired testing with clinical response as response variable

Combined timepoints, all arms only

```{r, warning=FALSE, echo=FALSE}
source('R/GSEA_plotting.R')

plot_cr_gsea_res(ptitle = 'unpaired',
                 bl_fn = 'rds2/GSEA_unpaired_baseline_TMM.rds',
                 pi_fn = 'rds2/GSEA_unpaired_postinduction_TMM.rds')
```

```{r, warning=FALSE, echo=FALSE}
source('R/GSEA_plotting.R')
graphics.off()

plot_cr_gsea_res(ptitle = 'unpaired_senescence',
                 bl_fn = 'rds2/GSEA_unpaired_baseline_TMM_senescence.rds',
                 pi_fn = 'rds2/GSEA_unpaired_postinduction_TMM_senescence.rds')
```



## Baseline

```{r, devel, eval = F, warning=FALSE, echo=FALSE}
source('R/GSEA_funcs_paired.R')
res <- gsea_wrapper(gene_sets = HALLMARK_pathways[1:3],
                    patients = rna_sample_annotation[arm == 'No induction',
                                                     unique(patient)],
                    resp_exp = 'timepoint',
                    exp_mat = rna_read_counts_salmon_tmm,
                    nperm = 10,
                    abs = F,
                    fn_extra = '')
```

```{r, warning=FALSE, echo=FALSE}
res <- readRDS(file = 'rds2/GSEA_unpaired_baseline.rds')
plot_es(res, legend.position = 'bottom', ptitle = 'unpaired',
        subset_var = NULL, x_var = 'arm')
```

### Heatmaps for leading edge genes.

```{r, warning=FALSE, echo=FALSE}
plot_leading_edge_heatmap(res, gsea_name = 'unpaired',
                          timepoint = timepoints[1],
                          distfun = 'euclidean',
                          fdr_thresh = .25)
```

### Parallel coordinates for leading edge genes.

```{r, warning=FALSE, echo=FALSE}
source('R/rna.R')
res <- readRDS(file = 'rds2/GSEA_unpaired_baseline.rds')
dtf <- compute_leading_edge_gene_scores(res, fdr_thresh = .25)
plot_leading_edge_gene_scores(dtf, allow_nonsensical = T,
                              res_name = 'BL_r_vs_nr_leading_edge')
```

## Post-induction

```{r, devel, eval = F, warning=FALSE, echo=FALSE}
source('R/GSEA_funcs_paired.R')
res <- gsea_all_arms(gene_sets = HALLMARK_pathways,
                     patients = rna_sample_annotation[!is.na(clinical_response),
                                                      unique(patient)],
                     allowed_timepoints = 'Post-induction',
                     gene_score_fn = my_unpaired_WC_test,
                     resp_exp = 'clinical_response',
                     paired_test = F,
                     nperm = 1e1,
                     abs = F,
                     fn_extra = '')
```

```{r, warning=FALSE, echo=FALSE}
res <- readRDS(file = 'rds2/GSEA_unpaired_postinduction.rds')
plot_es(res, legend.position = 'bottom', ptitle = 'unpaired',
        subset_var = NULL, x_var = 'arm')
```

```{r, warning=FALSE, echo=FALSE}
plot_leading_edge_heatmap(res, gsea_name = 'unpaired',
                          timepoint = timepoints[2],
                          distfun = 'euclidean',
                          fdr_thresh = .25)
```

### Parallel coordinates of leading edge gene set scores

```{r, warning=FALSE, echo=FALSE}
# source('R/load_rna_dat.R')
library(GSEAgenesets)
source('R/rna.R')
dtf <- compute_leading_edge_gene_scores(res, fdr_thresh = .25)
```

```{r, warning=FALSE, echo=FALSE}
source('R/rna.R')
plot_leading_edge_gene_scores(dtf, allow_nonsensical = T,
                              res_name = 'PI_r_vs_nr_leading_edge')
```

# CA15.3 transcriptomic associations

```{r, warning=FALSE, echo=FALSE}
source('~/TONIC/R/init.R')
source('R/load_rna_dat.R')
```

```{r, warning=FALSE, echo=FALSE}
readRDS(file = 'rds2/GSEA_ca15_3_baseline_TMM.rds') %>%
  dplyr::mutate(GeneSet = gsub('^HALLMARK_', '', GeneSet)) %>%
  plot_es(legend.position = 'right', ptitle = 'ca15_3',
          subset_var = NULL, x_var = 'arm')
```

```{r, warning=FALSE, echo=FALSE}
source('R/rna.R')

rna_read_counts_salmon <-
  file.path(p_root, 'salmon_rna', 'salmon_count_mat_gene_id.tsv') %>%
  fread %>% 
  normalize_colnames %>%
  as.data.frame %>%
  dplyr::mutate(gene_id = gsub('\\.\\d*', '', gene_id)) %>%
  column_to_rownames('gene_id') %>%
  as.matrix

contrast_ca15_3 <- function(l_timepoint = 'Baseline') {
  rna_subs <- rna_sample_annotation[timepoint == l_timepoint] %>%
    controlled_merge(patient_labels[, .(patient, timepoint, ca15_3)]) %>%
    mutate(ca15_3_bin = ca15_3 <= 
           patient_labels[!is.na(ca15_3) & clinical_response == 'R', 
                          max(ca15_3)])
  rna_subs <- rna_subs[!is.na(ca15_3)] %>% unique

  model_mat <- model.matrix(~ ca15_3_bin, rna_subs)
  colnames(model_mat) <- gsub('ca15_3_binTRUE', 'ca15_3_bin', colnames(model_mat))
  colnames(model_mat) <- gsub(':', '_', colnames(model_mat))
  colnames(model_mat) <- gsub('\\(Intercept\\)', 'intercept',
                              colnames(model_mat))
  print(colnames(model_mat))
  print(nrow(model_mat))
  # rownames(model_mat) <- rna_subs[, cf_number]

  idx  <- match(rna_subs$cf_number, colnames(rna_read_counts_salmon))
  cpms <- rna_read_counts_salmon[, idx]
  stopifnot(all(colnames(cpms) == rna_subs$cf_number))
  rownames(cpms) <- rownames(rna_read_counts_salmon)
  cpms <- cpms[!eps(apply(cpms, 1, var), 0, 1e-1), ]
  cpms <- limma::voom(cpms, design = model_mat)

  model_fit <- lmFit(cpms, design = model_mat)
  e_fit <- eBayes(model_fit, trend = F)
  return(e_fit)
}

bl_efit <- contrast_ca15_3(l_timepoint = 'Baseline')
# pi_efit <- contrast_ca15_3(l_timepoint = 'Post-induction')
# contrast_response(l_timepoint = 'Post-induction')
my_tt(bl_efit, coef = 2, sort.by = 'p', number = 100)$gene_symbol %>% writeLines
```

```{r}
source('R/rna.R')
my_tt(bl_efit, coef = 2, sort.by = 'T', resort.by = 'logFC', number = 1000)
my_tt(bl_efit, coef = 2, sort.by = 'p', number = 100)$gene_symbol %>% writeLines
```

```{r}
prep_gsea(patients = patient_labels[!is.na(ca15_3) & 
                                    timepoint == 'Baseline',
                                    patient],
          resp_exp = 'baseline_ca15_3_bin') 
```

# STING pathway activation

```{r, warning=FALSE, echo=FALSE}
source('~/TONIC/R/init.R')
source('R/load_rna_dat.R')
```

```{r, warning=FALSE, echo=FALSE}
source('R/GSEA_plotting.R')
rbind(
  cbind(readRDS(file = 'rds2/GSEA_unpaired_baseline_STING.rds')[[1]],
        timepoint = 'Baseline'),
  cbind(readRDS(file = 'rds2/GSEA_unpaired_postinduction_STING.rds')[[1]],
        timepoint = 'Post-induction')) %>%
  dplyr::mutate(GeneSet = gsub('^HALLMARK_', '', GeneSet)) %>%
  plot_es(x_var = 'timepoint', ptitle = 'STING')
plot_cr_gsea_res(ptitle = 'STING', 
                 bl_fn = 'rds2/GSEA_unpaired_baseline_STING.rds'
                 pi_fn = 'rds2/GSEA_unpaired_postinduction_STING.rds'
```

```{r, warning=FALSE, echo=FALSE}
hist(readRDS(file = 'rds2/GSEA_unpaired_baseline_STING.rds')[[2]])
hist(readRDS(file = 'rds2/GSEA_unpaired_postinduction_STING.rds')[[2]])
readRDS(file = 'rds2/GSEA_unpaired_postinduction_STING.rds')[[2]] %>%
  { .[abs(.) >= .999] } %>%
  { .[order(abs(.))] } 

# plot_es(p_dat, legend.position = 'right', ptitle = 'paired_senescence',
#         subset_var = NULL)

p <- readRDS(file = 'rds2/GSEA_unpaired_postinduction_STING.rds')[[1]] %>%
  plot_es(legend.position = 'right', ptitle = 'paired_senescence',
          subset_var = NULL)
```


```{r, warning=FALSE, echo=FALSE}
devtools::load_all('~/libs/GSEAgenesets')

plot_GSEA_scores(fn = 'rds2/GSEA_unpaired_postinduction_STING.rds')
plot_GSEA_scores(fn = 'rds2/GSEA_unpaired_baseline_STING.rds')
```
