---
title: "TIL and blood TCR repertoires summary statistics"
author: "Maarten Slagter"
output:
  html_document:
    toc: true
    number_sections: true
    theme: united
---


```{r, warning=FALSE, echo=FALSE, cache = F}
knitr::opts_chunk$set(message = FALSE, cache = T, cache.lazy = F,
                      cache.comments = F, autodep = T, warning = FALSE,
                      results = 'hide', fig.keep = 'high',
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# knitr::clean_cache()
# knitr::opts_chunk$get("cache.path")
# knitr::knit_global()
setwd('~/TONIC')
source(file.path('R', 'init.R'))
# read_adaptive_seqs()
```


# Single timepoint summary statistic associations

Repertoire size is almost correlated with response at
baseline, but not post-induction.
On the contrary, clonality is associated with response post-induction and on
nivo, T-cell infiltration only on-nivo.

## Tumor

```{r, warning=FALSE, echo=FALSE, results='asis'}
res <- test_adaptive_association(timepoint = 'Baseline')
print(knitr::kable(res))
res <- test_adaptive_association(timepoint = 'Post-induction')
print(knitr::kable(res))
res <- test_adaptive_association(timepoint = 'On nivo')
print(knitr::kable(res))
```

Visualisations of the comparisons shown above in table form

```{r, warning=FALSE, echo=FALSE, fig.width = 3, fig.height = 4}
# source('R/plotting_adaptive.R')
# clear_object('as_plots_bl', sr)
# clear_object('as_plots_pi', sr)
# clear_object("as_plots_on", sr)
sr(as_plots_bl <- plot_single_timepoint_r_assocation(l_timepoint = 'Baseline',
                                   facet_var = NULL))
print(as_plots_bl)
sr(as_plots_pi <- plot_single_timepoint_r_assocation(l_timepoint = 'Post-induction',
                                   facet_var = NULL))
print(as_plots_pi)
sr(as_plots_on <- plot_single_timepoint_r_assocation(l_timepoint = 'On nivo',
                                   facet_var = NULL))
print(as_plots_on)
```

The same comparisons but split up by induction arm

```{r, warning=FALSE, echo=FALSE, fig.width = 5, fig.height = 4}
# source('R/plotting_adaptive.R')
plot_single_timepoint_r_assocation(l_timepoint = 'Baseline',
                                   facet_var = 'arm')
plot_single_timepoint_r_assocation(l_timepoint = 'Post-induction',
                                   facet_var = 'arm')
plot_single_timepoint_r_assocation(l_timepoint = 'On nivo',
                                   facet_var = 'arm')
```

We modelled clinical response with the ImmunoSeq summary statistics (clonality,
T cell infiltration and repertoire size as estimated using the Efron-Thisted
method) and evaluated model performance in 10-fold cross validation.  This
further confirmed that intatumoral T cell compartment was informative at the
post-induction and on nivolumab timepoints, whereas the summary statistics of
the peripheral T cell compartment weren't informative at all.


```{r, warning=FALSE, echo=FALSE, results='asis'}
pacman::p_load(glmnet)
measures = c('sample_clonality', 'efron_thisted_estimator',
             'adaptive_t_cells')

lbs_fun <- function(fit, ...) {
  L <- length(fit$lambda)
  x <- log(fit$lambda[L])
  y <- fit$beta[, L]
  text(x, y, labels = names(y), adj = 0, ...)
}

plot_cv_glmnet <- function(tp, modelversion, cv.glmnet, predicted) {
  layout(c(1, 2, 3))

  plot(cv.glmnet, main = sprintf('%s %s', tp, modelversion),
       xlab = 'log(Lambda)')
  # abline(a = null_performance, b = 0, col = 'red', lwd = 2)
  null_performance <-
  print(coef(cv.glmnet, s = "lambda.min"))

  plot(cv.glmnet$glmnet.fit, label = F, xvar = 'lambda', xlab = 'log(Lambda)')
  abline(v = log(cv.glmnet$lambda.min), lty =3)
  abline(v = log(cv.glmnet$lambda.1se), lty =3)
  lbs_fun(cv.glmnet$glmnet.fit)

  boxplot(predicted$actual, predicted$predicted)
}

rank_percentile <- function(v) {
  (frank(v, ties.method = 'max') - 1) / (length(v) - 1)
}

range_transform <- function(v) {
  (v - min(v)) / (max(v) - min(v))
}
# c(1, 4, 6, 5) %>% range_transform
# c(1, 4, 6, 5) %>% rank_percentile

norm_method = 'none'
norm_method = 'Z'
alpha = 0
plot_glmnet_analyses <- function(norm_method = 'range', alpha = 1,
                                 iterative_plotting = F) {
  for (tp in c(timepoints, blood_timepoints)) {
    if (tp %in% timepoints) {
      t_dat <- patient_labels[timepoint == tp,
                              c(measures, 'clinical_response'), with = F]
    } else {
      t_dat <- blood_adaptive[blood_timepoint == tp,
                              c(measures, 'clinical_response'), with = F]
    }
    t_dat %<>% { .[complete.cases(.)] }
    if (t_dat[, .N] <= 10) next

    ## Single covariates
    X_mat <- as.matrix(t_dat[, 1:3])
    if (norm_method == 'Z') {
      X_mat <- apply(X_mat, 2, function(x) (x - mean(x)) / sd(x))
    } else if (norm_method == 'rank') {
      X_mat <- apply(X_mat, 2, rank_percentile)
    } else if (norm_method == 'range') {
      X_mat <- apply(X_mat, 2, range_transform)
    }
    X_mat <- apply(X_mat, 2, function(x) (x - mean(x)) / sd(x))
    y_vec <- as.integer(t_dat[, 4] == 'R')
    null_performance <- mean(y_vec) %>% { ifelse(. > .5, 1 - ., .) }
    cv.glmnet <- cv.glmnet(x = X_mat, y = y_vec, family = 'binomial',
                           alpha = alpha)
    predicted <- data.frame('actual' = as.factor(y_vec),
                            'predicted' = predict(cv.glmnet, newx = X_mat,
                                                  s = 'lambda.min')[, 1])
    predicted %>%
      group_by(actual) %>%
      dplyr::summarise(median(predicted)) %>%
      print

    plot_cv_glmnet(tp, 'simple', cv.glmnet, predicted)
    if (iterative_plotting && interactive())
      invisible(readline(prompt = 'Press [enter] to continue'))

    ## Add interaction terms
    # X_mat <- as.matrix(t_dat[, 1:3])
    iter <- 1
    for (i in seq(1, length(measures) - 1)) {
      for (j in seq(i + 1, length(measures))) {
        X_mat %<>% cbind(X_mat[, i] * X_mat[, j])
        colnames(X_mat)[3 + iter] <- sprintf('%s.%s', measures[i], measures[j])
        iter <- iter + 1
      }
    }
    if (norm_method == 'Z') {
      X_mat <- apply(X_mat, 2, function(x) (x - mean(x)) / sd(x))
    } else if (norm_method == 'rank') {
      X_mat <- apply(X_mat, 2, rank_percentile)
    } else if (norm_method == 'range') {
      X_mat <- apply(X_mat, 2, range_transform)
    }
    cv.glmnet <- cv.glmnet(x = X_mat, y = y_vec, family = 'binomial',
                           alpha = alpha)
    plot_cv_glmnet(tp, 'simple', cv.glmnet,
                   data.frame('actual' = as.factor(y_vec),
                              'predicted' = predict(cv.glmnet, newx = X_mat,
                                                    s = 'lambda.min')[, 1]))
    if (iterative_plotting && interactive())
      invisible(readline(prompt = 'Press [enter] to continue'))
  }
}

pdf(file.path(img_dir, 'logistic_regression_adaptive_summary_statistics.pdf'),
    width = 7, height = 10)
plot_glmnet_analyses()
dev.off()

# data.frame('actual' = as.factor(y_vec),
#            'predicted' = predict(cv.glmnet, newx = X_mat, s = 'lambda.min')[, 1]) %>%
#   ggplot(aes(x = actual, y = predicted)) + geom_boxplot()
# cv.glmnet
```

###

## Blood

```{r, warning=FALSE, echo=FALSE, results='asis'}
res <- test_adaptive_association(timepoint = '-2')
print(knitr::kable(res))
res <- test_adaptive_association(timepoint = '0')
print(knitr::kable(res))
res <- test_adaptive_association(timepoint = '6')
print(knitr::kable(res))
# res <- test_adaptive_association(timepoint = '12')
# print(knitr::kable(res))
```


```{r, warning=FALSE, echo=FALSE, fig.width = 3, fig.height = 4}
sr(as_plots_blood_bl <- plot_single_timepoint_r_assocation(l_timepoint = '-2',
                                   facet_var = NULL))
# print(as_plots_blood_bl)
sr(as_plots_blood_pi <- plot_single_timepoint_r_assocation(l_timepoint = '0',
                                   facet_var = NULL))
# print(as_plots_blood_pi)
sr(as_plots_blood_on <- plot_single_timepoint_r_assocation(l_timepoint = '6',
                                   facet_var = NULL))
# print(as_plots_blood_on)
```

```{r, warning=FALSE, echo=FALSE, fig.width = 5, fig.height = 4}
plot_single_timepoint_r_assocation(l_timepoint = '-2',
                                   facet_var = 'arm')
plot_single_timepoint_r_assocation(l_timepoint = '0',
                                   facet_var = 'arm')
plot_single_timepoint_r_assocation(l_timepoint = '6',
                                   facet_var = 'arm')
```

## Plot all arm-unspecific summary stats into one overview

```{r, eval = F, warning=FALSE, echo=FALSE, fig.width = 5, fig.height = 4}
# devtools::load_all('~/libs/maartenutils')
plot_panel_layout(plots = c(as_plots_bl, as_plots_pi, as_plots_on,
                            as_plots_blood_bl, as_plots_blood_pi,
                            as_plots_blood_on),
                  ncol = 3, nrow = 3,
                  labels = c(tolower(LETTERS), paste0(tolower(LETTERS), "'")),
                  filename = file.path(img_dir, 'as_summary_stats.pdf'))
```


# Relationship between distribution and induction treatment & clinical response

Bivariate spider plots

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_adaptive.R')
# source('R/plotting_nanostring.R')
sr(comb_time_resp_palette)

plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'Post-induction', 'On nivo'),
                 plot_counts = T,
                 var1 = 'sample_clonality',
                 var2 = 'adaptive_t_cells',
                 colour_var = 'comb_time_resp')

ggsave(filename = file.path(img_dir,
                            'adaptive_sum_stats_spider_plot.pdf'),
       width = 17.4, height = 11, units = 'cm')
```

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_adaptive.R')
# source('R/plotting_nanostring.R')
sr(comb_time_resp_palette)

plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'Post-induction'),
                 plot_counts = T,
                 var1 = 'sample_clonality',
                 var2 = 'adaptive_t_cells',
                 colour_var = 'comb_time_resp')

ggsave(filename = file.path(img_dir,
                            'adaptive_sum_stats_spider_plot_pi.pdf'),
       width = 17.4, height = 11, units = 'cm')
```

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_adaptive.R')
# source('R/plotting_nanostring.R')
sr(comb_time_resp_palette)

plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'On nivo'),
                 plot_counts = T,
                 var1 = 'sample_clonality',
                 var2 = 'adaptive_t_cells',
                 colour_var = 'comb_time_resp')

ggsave(filename = file.path(img_dir,
                            'adaptive_sum_stats_spider_plot_on.pdf'),
       width = 17.4, height = 11, units = 'cm')
```

The same but for PBL repertoires

```{r, warning=FALSE, echo=FALSE}
# source('R/plotting_adaptive.R')
# source('R/plotting_nanostring.R')

if ('blood_timepoint' %in% colnames(blood_adaptive)) {
  setnames(blood_adaptive, 'blood_timepoint', 'timepoint')
}

plot_adaptive_FC(dtf = blood_adaptive,
                 facet_var = 'arm',
                 p_timepoints = blood_timepoints,
                 var1 = 'sample_clonality',
                 var2 = 'adaptive_t_cells',
                 colour_var = 'clinical_response')
```

```{r, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'Post-induction'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'comb_time_resp')
```

```{r, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'comb_time_resp')
```

The same, but not coloring by clinical response

```{r, warning=FALSE, echo=FALSE}
# source('R/plotting_adaptive.R')
plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'Post-induction', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'timepoint')
```

```{r, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'Post-induction'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'timepoint')
```

```{r, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'timepoint')
```

The same, but not coloring by clinical response and showing %TIL instead

```{r, spider1, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'Post-induction', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'adaptive_t_cells',
                 colour_var = 'timepoint')
```

```{r, spider2, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'Post-induction'),
                 var1 = 'sample_clonality', var2 = 'adaptive_t_cells',
                 colour_var = 'timepoint')
```

```{r, spider3, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm',
                 p_timepoints = c('Baseline', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'adaptive_t_cells',
                 colour_var = 'timepoint')
```

```{r, warning=FALSE, echo=FALSE, fig.width = 4, fig.height = 6}
# source('R/plotting_nanostring.R')
# source('R/plotting_adaptive.R')
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'tumor',
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'tumor',
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'tumor',
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'blood',
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'blood',
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'blood',
                       facet_var = NULL)
```

The same but restricting to some time points for clarity: -2 (pre) and 6 (on
nivo)

```{r, warning=FALSE, echo=FALSE, fig.width = 4, fig.height = 6}
# source('R/plotting_nanostring.R')
# source('R/plotting_adaptive.R')
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'blood',
                       facet_var = NULL,
                       x_axis_tps = c(-2, 6))
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'blood',
                       facet_var = NULL,
                       x_axis_tps = c(-2, 6))
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'blood',
                       facet_var = NULL,
                       x_axis_tps = c(-2, 6))
```

```{r, fig.width = 7, fig.height = 3, warning=FALSE, echo=FALSE}
# source('R/plotting_nanostring.R')
# source('R/plotting_adaptive.R')
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'tumor',
                       facet_var = 'arm')
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'tumor',
                       facet_var = 'arm')
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'tumor',
                       facet_var = 'arm')
```

```{r, fig.width = 7, fig.height = 3, warning=FALSE, echo=FALSE}
# debugonce(plot_parallel_adaptive)
plot_parallel_adaptive(p_var = 'efron_thisted_estimator',
                       compartment = 'blood',
                       facet_var = 'arm')
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'blood',
                       facet_var = 'arm')
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'blood',
                       facet_var = 'arm')
```

```{r, fig.width = 7, fig.height = 3, warning=FALSE, echo=FALSE}
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'blood',
                       facet_var = 'arm',
                       x_axis_tps = c(-2, 6))
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'blood',
                       facet_var = 'arm',
                       x_axis_tps = c(-2, 6))
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'blood',
                       facet_var = 'arm',
                       x_axis_tps = c(-2, 6))
```

# Deltas in summary statistics over time

```{r, warning=FALSE, echo=FALSE}
# source('R/plotting_nanostring.R')
source('R/misc.R')
source('R/plotting_adaptive.R')
plots <- plot_all_comps(f = plot_adaptive_summary_stats,
                        colour_var = 'clinical_response',
                        x_var = 'clinical_response',
                        facet_var = NULL)
plot_panel_layout(plots, ncol = 2,
                  labels = NULL,
                  filename = file.path(img_dir,
                                       'adaptive_summary_stat_deltas.pdf'))

t_dats <- plot_all_comps(f = compare_adaptive_summary_stats,
                         colour_var = 'clinical_response',
                         x_var = 'clinical_response',
                         facet_var = NULL)
write_tsv(purrr::imap_dfr(t_dats, function(x, y) cbind(x, 'analysis_name' = y)),
          file.path(p_root, 'ext', 'adaptive_summary_stat_deltas.tsv'))
```

Effect of induction arm on these measures

```{r, eval = T, warning=FALSE, echo=FALSE}
# graphics.off()
# source('R/plotting_nanostring.R')
# source('R/plotting_adaptive.R')
plots <- plot_all_comps(f = plot_adaptive_summary_stats,
                        colour_var = 'arm',
                        x_var = 'arm',
                        facet_var = NULL)
plot_panel_layout(plots, ncol = 1,
                  nrow = 3,
                  labels = NULL,
                  filename = file.path(img_dir,
                                       'adaptive_summary_stat_deltas_by_arm.pdf'))

t_dats <- plot_all_comps(f = compare_adaptive_summary_stats,
                         colour_var = 'arm',
                         x_var = 'arm',
                         facet_var = NULL)
write_tsv(purrr::imap_dfr(t_dats, function(x, y)
                          cbind(x, 'analysis_name' = y)),
          file.path(p_root, 'ext', 'adaptive_summary_stat_deltas_by_arm.tsv'))
```

```{r,  warning=FALSE, echo=FALSE}
graphics.off()
# source('R/plotting_adaptive.R')
compare_adaptive_summary_stats(comp_measure = 'sample_clonality',
                               facet_var = NULL)
```

Split out by clinical response, contrasts most pronounced in responders

```{r, warning=FALSE, echo=FALSE}
# source('R/plotting_nanostring.R')
# source('R/plotting_adaptive.R')
plot_all_comps(f = compare_adaptive_summary_stats,
               colour_var = 'clinical_response',
               x_var = 'clinical_response',
               facet_var = 'arm')
```

