---
title: "Potential confounding variables in transcriptomic analyses"
author: "Maarten Slagter"
output: html_document
---

```{r, cache = F, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = F, cache.lazy = F, 
                      cache.comments = F, autodep = T, warning = FALSE, 
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
```

The relatively small patient numbers per induction arm render the TONIC study
susceptible to potentially confounding variables not being equal between
treatment arms. In an initial clinical response analysis, I already observed
some clinical variables like WHO PS and CA15.3 are unevenly distributed between
treatment arms and correcting for them changed the relative order given to
different induction arms. In the document, I investigate whether we should
similarly control for baseline covariates in NanoString signature fold change
analysis. Do potential confounders increase the likelihood of showing a large
fold change?

```{r, init, eval = T, warning=FALSE, echo=FALSE}
source('R/init.R')
## Read data
#{
patient_labels <- read.csv(file.path(p_root, 'data-raw',
                                     '180212_clinical_data_all_patients.csv'),
                           dec = ',', sep = ';') %>% as.data.table %>%
  normalize_colnames()
setnames(patient_labels, gsub('\\.', '_', colnames(patient_labels)))
maartenutils::set_dt_types(patient_labels,
                           c('mean_log2_hk' = 'numeric',
                             'tis_score' = 'numeric'))
patient_labels[, patient := paste0('pat_', study_id)]
patient_labels[is.na(patient)]
patient_labels[, table(patient)]
patient_labels <- patient_labels[1:67]
names(patient_labels)
base::grep(x = names(patient_labels), pattern = 'til')
patient_labels[, 'performance_status' := droplevels(performance_status)]
patient_labels[, table(performance_status)]

patient_labels[, .(patient, induction_therapy, performance_status,
                   response_rate, pd_l1_immuuninfiltrate)]
colnames(patient_labels)
# patient_labels[patient == 'pat_31']
## Exclude patient 26, the MMR patient
patient_labels <- patient_labels[-which.max(time_progression_random_weeks)]
# patient_labels[which.max(time_progression_random__weeks)]

# clear_object('brm_tonic', sr)
# clear_all_objects(sr = sr)
```


```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
```

```{r, warning=FALSE, echo=FALSE}
test_confounding <- function(x_var = 'Baseline',
                             x_label = 'Baseline expression level',
                             tp1 = 'Baseline', tp2 = 'Post-induction',
                             feature = 'tis') {
  c_dat_exp <- prepare_nano_bivariates(x_var = tp1,
                                       y_var = tp2,
                                       colour_var = 'arm') %>%
    merge(clinical_annotation, by = 'patient')

  ggplot(c_dat_exp[variable == feature], aes_string(x = x_var, y = "logfc")) +
    geom_point() +
    facet_grid(~arm) +
    geom_smooth(se = F, span = 2, alpha = .2) +
    xlab(x_label) +
    ylab(sprintf('log FC\n%s vs. %s', tp1, tp2)) +
    ggtitle(feature)
}

plot_confounding_panel <- function(tp1 = 'Baseline', tp2 = 'Post-induction',
                                   feature = 'tis') {
  cowplot::plot_grid(
    test_confounding(x_var = sprintf('`%s`', tp1),
                     x_label = sprintf('%s expression level', tp1),
                     tp1 = tp1, tp2 = tp2, feature = feature),
    test_confounding(x_var = 'ca15_3', x_label = 'Baseline CA15.3',
                     tp1 = tp1, tp2 = tp2, feature = feature),
    test_confounding(x_var = 's_til', x_label = 'Baseline stromal TIL [%]',
                     tp1 = tp1, tp2 = tp2, feature = feature),
    test_confounding(x_var = 'pd_l1_immunoinfiltrate',
                     x_label = 'Baseline PD-L1 [%]',
                     tp1 = tp1, tp2 = tp2, feature = feature),
    ncol = 1)
}

# plot_confounding_panel(tp1 = 'Baseline', tp2 = 'Post-induction',
#                        feature = 'tis')
# plot_confounding_panel(tp1 = 'Baseline', tp2 = 'On nivo',
#                        feature = 'tis')
# plot_confounding_panel(tp1 = 'Post-induction', tp2 = 'On nivo',
#                        feature = 'tis')
```

```{r, warning=FALSE, echo=FALSE}
test_confounding_heatmap <- function(tp1 = 'Baseline', tp2 = 'Post-induction') {
  c_dat_exp <- prepare_nano_bivariates(x_var = tp1,
                                       y_var = tp2,
                                       colour_var = 'arm') %>%
    merge(clinical_annotation, by = 'patient')
  # x_label = sprintf('%s expression level', tp1)
  # x_label = 'Baseline PD-L1 [%]'
  # comp_vars <- c(sprintf('`%s`', tp1), 'ca15_3', 's_til',
  #                'pd_l1_immunoinfiltrate')
  comp_vars <- c(tp1, 'ca15_3', 's_til', 'pd_l1_immunoinfiltrate')

  plots <- lapply(c_dat_exp[, unique(variable)], function(feat) {
    p_dat <- rbindlist(lapply(auto_name(comp_vars), function(comp_var) {
        c_dat_exp[variable == feat,
                  .('cor' = cor(logfc, get(comp_var), method = 'spearman'),
                    'var' = comp_var,
                    'feat' = feat),
                  by = arm]
        }), fill = T)

    ggplot(p_dat, aes(x = arm, y = var, fill = cor,
                      label = sprintf('%.2f', cor))) +
      geom_raster() +
      geom_text(size = 2) +
      scale_fill_gradient2(name = 'Spearman correlation',
                           low = scales::muted('red'),
                           high = scales::muted('blue'),
                           limits = c(-1, 1),
                           na.value = 'white') +
      ggtitle(feat) +
      rotate_x_labels(45) +
      xlab('') + ylab('')
  })

  plot_panel_layout(plots, ncol = 2, nrow = 3,
                    labels = NULL,
                    clear_redundant_labels = T,
                    clear_redundant_legends = T,
                    filename = file.path(plot_dir, 'baseline_logfc_cors.pdf'))
}
test_confounding_heatmap()
```

A gene signature's baseline levels anti-correlate with the FC that's achieved by
all induction therapies. This can probably be understood by the log
transformation that is applied to the gene signature scores. This results in
increasing difficulty for a signature score to increase with high baseline
levels. Shown below are all correlation coefficients for all of the gene
signatures fold changes with the four indicated variables in the panel rows.

```{r, warning=FALSE, echo=FALSE}
test_confounding_cor_hists <- function(tp1 = 'Baseline', tp2 = 'Post-induction') {
  c_dat_exp <- prepare_nano_bivariates(x_var = tp1,
                                       y_var = tp2,
                                       colour_var = 'arm') %>%
    merge(clinical_annotation, by = 'patient')

  comp_vars <- c(tp1, 'ca15_3', 's_til', 'pd_l1_immunoinfiltrate')

  t_dat <- rbindlist(lapply(c_dat_exp[, unique(variable)], function(feat) {
    rbindlist(lapply(auto_name(comp_vars), function(comp_var) {
      c_dat_exp[variable == feat,
                .('cor' = cor(logfc, get(comp_var), method = 'spearman'),
                  'var' = comp_var,
                  'feat' = feat),
                by = arm]
      }), fill = T) 
  }), fill = T)

  t_dat[, .('frac_pos' = .SD[cor > 0, .N] / .SD[!is.na(cor), .N]), 
        by = c('var', 'arm')]
  
  ggplot(t_dat, aes(x = cor)) + 
    facet_grid( var ~ arm, labeller = label_value) +
    geom_vline(xintercept = 0, colour = 'grey85', type = 'dashed') +
    geom_histogram() + 
    xlab('Spearman correlation coefficients') +
    ylab('Number of sigs') +
    geom_vline(data = t_dat[, .('median' = median(cor, na.rm = T)), 
                            by = c('var', 'arm')],
               aes(xintercept = median), colour = 'red')

}

test_confounding_cor_hists()
```

# Formal tests

```{r, warning=FALSE, echo=FALSE}
# clear_object(tis_baseline_exp)
sr(tis_baseline_exp <- do.call('brm',
                               c(list(formula = bf(logfc ~ 0 + (1 | arm)),
                                      data = test_dat,
                                      prior = NULL),
                                 list('family' = gaussian(),
                                      'warmup' = 1000,
                                      'iter' = 3e3,
                                      'chains' = 4,
                                      'control' = list(adapt_delta = 0.95)))))
# stancode(tis_baseline_exp)
update_model <- function(fit = tis_baseline_exp, feat = 'tis',
                         newdata = c_dat_exp[variable == feat]) {
  update(fit, newdata = newdata)
}
# update_model(feat = 'apm')
```
```{r, warning=FALSE, echo=FALSE}
stanplot(tis_baseline_exp, pars = '^r_.*(arm)', type = 'intervals')
updated_model <- update_model(feat = 'apm')
stanplot(updated_model, pars = '^r_.*(arm)', type = 'intervals')
```

Expand with baseline CA15.3, this does not influence fold changes at all

```{r, warning=FALSE, echo=FALSE}
sr(tis_baseline_exp_e1 <- do.call('brm',
                               c(list(formula = bf(logfc ~ 0 + (1 + ca15_3 | arm)),
                                      data = test_dat,
                                      prior = NULL),
                                 list('family' = gaussian(),
                                      'warmup' = 1000,
                                      'iter' = 3e3,
                                      'chains' = 4,
                                      'control' = list(adapt_delta = 0.95)))))
```

Also not in TIS

```{r, warning=FALSE, echo=FALSE}
stanplot(tis_baseline_exp_e1, pars = '^r_.*(arm)', type = 'intervals')
updated_model <- update_model(fit = tis_baseline_exp_e1, feat = 'tis')
stanplot(updated_model, pars = '^r_.*(arm)', type = 'intervals')
```

Expand with baseline CA15.3, this does not influence fold changes at all

```{r, warning=FALSE, echo=FALSE}
sr(tis_baseline_exp_e2 <- do.call('brm',
                               c(list(formula = bf(logfc ~ 0 + (1 + s_til | arm)),
                                      data = test_dat,
                                      prior = NULL),
                                 list('family' = gaussian(),
                                      'warmup' = 1000,
                                      'iter' = 3e3,
                                      'chains' = 4,
                                      'control' = list(adapt_delta = 0.95)))))
```


# Control for baseline covariates in clinical response comparison

```{r, warning=FALSE, echo=FALSE}
priors <- NULL
priors <- c(#set_prior("beta(1,1)", class = "b", coef = "response_bin"),
            set_prior("cauchy(1,2)", class = "b"))
```

```{r, warning=FALSE, echo=FALSE}
sr(brm_tonic <- do.call('brm',
                     c(list(formula = bf(response_bin ~ (1 | induction_therapy) +
             who_performance_status + s_til + tis_nanostring + ca15_3 + age,
             nl = F), data = clinical_annotation, prior = NULL),
                       brms_settings)))
dat <- clinical_annotation[study_id %in% as.integer(rownames(standata(brm_tonic)$X))]
```

```{r, warning=FALSE, echo=FALSE}
sr(brm_tonic_basic <- do.call('brm',
  c(list(formula = bf(response_bin ~ (1 | induction_therapy), nl = F),
         data = dat,
         prior = NULL),
    brms_settings)))
```

```{r, warning=FALSE, echo=FALSE}
sr(brm_tonic_intermed <- do.call('brm',
  c(list(formula = bf(response_bin ~ (1 | induction_therapy) + who_performance_status,
                      nl = F),
         data = dat,
         prior = NULL),
    brms_settings)))
```

```{r, warning=FALSE, echo=FALSE}
sr(brm_tonic_ca15 <- do.call('brm',
  c(list(formula = bf(response_bin ~ (1 | induction_therapy) +
                  who_performance_status + ca15_3,
                  nl = F),
         data = dat,
         prior = NULL),
    brms_settings)))
```

```{r, warning=FALSE, echo=FALSE}
sr(brm_tonic_s_til <- do.call('brm',
  c(list(formula = bf(response_bin ~ (1 | induction_therapy) +
                  who_performance_status + s_til,
                  nl = F),
         data = dat,
         prior = NULL),
    brms_settings)))
```

```{r, warning=FALSE, echo=FALSE}
# source('R/init.R')
# brm_tonic_basic <- readRDS(file.path(rds_dir, 'brm_tonic_basic.rds'))
# brm_tonic_intermed <- readRDS(file.path(rds_dir, 'brm_tonic_intermed.rds'))
# brm_tonic_ca15 <- readRDS(file.path(rds_dir, 'brm_tonic_ca15.rds'))
# brm_tonic <- readRDS(file.path(rds_dir, 'brm_tonic.rds'))
# perform_model_comps(brms_object = brm_tonic_basic)
# perform_model_comps(brms_object = brm_tonic_intermed)
# perform_model_comps(brms_object = brm_tonic)
```

```{r, warning=FALSE, echo=FALSE}
# posterior_histogram(brms_object = brm_tonic_basic,
#                     name = 'Model: Induction therapy',
#                     variables = NULL)

p <- plot_panel_layout(list(
  posterior_histogram(brms_object = brm_tonic_basic,
                      name = 'Model: Induction therapy',
                      variables = c('Doxorubicin', 'Cisplatin', 'No.induction')) +
                       theme(legend.position = 'right'),
  # posterior_histogram(brms_object = brm_tonic_basic,
  #                     name = 'Model: Induction therapy',
  #                     offset_val = 2,
  #                     variables = c('Doxorubicin', 'Cisplatin', 'No.induction')),
  posterior_histogram(brms_object = brm_tonic_intermed,
                      name = 'Model: Induction therapy + WHO PS',
                      variables = c('Doxorubicin', 'Cisplatin', 'No.induction')),
  # posterior_histogram(brms_object = brm_tonic_intermed,
  #                     name = 'Model: Induction therapy + WHO PS',
  #                     offset_val = 2,
  #                     variables = c('Doxorubicin', 'Cisplatin', 'No.induction')),
  posterior_histogram(brms_object = brm_tonic_ca15,
                      name = 'Model: Induction therapy + WHO PS + CA15.3',
                      offset_val = 1,
                      variables = c('Doxorubicin', 'Cisplatin', 'No.induction'))),
                  ncol = 1, labels = NULL,
                  filename = file.path(plot_dir, 'bayesian_hists_no_MMR.png'),
                  w = 10, h = 15)
  # posterior_histogram(brms_object = brm_tonic_ca15,
  #                     name = 'Model: Induction therapy + WHO PS + CA15.3',
  #                     offset_val = 2,
  #                     variables = c('Doxorubicin', 'Cisplatin', 'No.induction'))),
```
