```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
source('R/load_rna_dat.r')
```

# Adaptive TCR-Seq correlations with clinical data

```{r, warning=FALSE, echo=FALSE}
ggplot(patient_labels, aes(x = clinical_response, y = observed_richness)) + 
  geom_boxplot() + facet_grid(arm ~ timepoint)
ggplot(patient_labels, aes(x = clinical_response, y = pielou_evenness)) + 
  geom_boxplot() + facet_grid(arm ~ timepoint)
```

Adaptive data over time

```{r, warning=FALSE, echo=FALSE}
# ggplot(patient_labels, aes(x = timepoint, y = observed_richness, 
ggplot(patient_labels, aes(x = timepoint, y = pielou_evenness, 
                           colour = clinical_response, group = patient)) +
  geom_point() +
  geom_line() +
  facet_grid(~ arm)
```

# NanoString correlations with clinical response

```{r, warning=FALSE, echo=FALSE}
timepoint = 'Baseline'
source('R/test_associations.R')
# lapply(danaher_signature_genes[, auto_name(unique(signature_name))],
#        function(sig) {
#          test_gene_set_association(gene_set = sig, timepoint = timepoint)
#        })
p_dat <- test_gene_set_association(
  gene_set = danaher_scores.m[, unique(variable)], timepoint = timepoint)
p_dat <- test_gene_set_association(
  gene_set = danaher_scores.m[, unique(variable)], timepoint = 'Post-induction')
p_dat <- test_gene_set_association(
  gene_set = danaher_scores.m[, unique(variable)], timepoint = 'On nivo')
# p_dat[p_val <= .05]
```

```{r, warning=FALSE, echo=FALSE}
plot_volcano <- function(gene_sets = danaher_scores.m[, unique(variable)],
                         timepoint = 'Baseline') {
  p_dat <- test_gene_set_association(gene_set = gene_sets,
                                     timepoint = timepoint)

  ggplot(p_dat, aes(x = log2FC, y = -log10(p_val), label = variable)) +
    geom_point() +
    ggrepel::geom_text_repel(data = p_dat[p_val <= .05]) +
    geom_hline(yintercept = -log10(.05),
               color = 'darkred', linetype = 'dashed') +
    geom_vline(xintercept = 0, color = 'grey70', linetype = 'dashed') +
    scale_y_continuous(name = 'Unadjusted p-value', trans = 'identity',
                       breaks = c(-seq(-5,0, by = .5), -log10(0.05)),
                       limits = c(0, 3.5),
                       labels = function(x)
                         parse(text = sprintf('10^{-%.2f}', x))) +
    ggtitle(timepoint) +
    xlab(sprintf('Fold difference\nresponders and non-responders (%s)',
                 timepoint))
}

plots <- list(
  plot_volcano(timepoint = 'Baseline'),
  # plot_volcano(timepoint = 'Post-induction'),
  plot_volcano(timepoint = 'On nivo'))
# cowplot::plot_grid(plots[[1]], plots[[2]], plots[[3]], ncol = 3)
ggsave(filename = file.path(img_dir, 'gene_sigs_response_volcanoes.pdf'),
       width = 29, height = 11, units = 'cm')
cowplot::plot_grid(plots[[1]], plots[[2]], ncol = 2)
ggsave(filename = file.path(img_dir, 'gene_sigs_response_volcanoes.pdf'),
       width = 24, height = 11, units = 'cm')
```


# Correlations between RNA and NanoString

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
plots <- lapply(danaher_signature_genes[, unique(signature_name)],
                function(sig) {
  compare_nano_rna_sigs(sig = sig)
}) %>% { .[!sapply(., is.null)] }

maartenutils::plot_panel_layout(plots, ncol = 3, nrow = 4,
  labels = NULL,
  plot_direct = F,
  ref_panel_idx = 3 * 3,
  clear_redundant_labels = T,
  filename = file.path(plot_dir, sprintf('rna_nano_cors_%s.pdf', colour_var)),
  w = 17.4, h = 25, label_size = 8)
```


## The same but for individual genes

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')

rna_nano_cors <- rbindlist(lapply(auto_name(unique(test_genes)), function(gs) {
  t_dat <- correlate_nano_rna(genes = gs)
  if (null_dat(t_dat)) return(NULL)
  cors <- t_dat[, .('pearson' = cor(rna_value, nano_value,
                                    method = 'pearson'),
                    'spearman' = cor(rna_value, nano_value,
                                     method = 'spearman'),
                    rna_q50, nano_q50),
                by = c('gene_symbol', 'timepoint')]
  return(cors)
}), fill = T)

rna_nano_cors_sigs <- rbindlist(lapply(auto_name(danaher_scores.m[, levels(variable)]),
                                  function(sig) {
  t_dat <- correlate_nano_rna(sig = sig, genes = NULL)
  if (null_dat(t_dat)) return(NULL)
  cors <- t_dat[, .('pearson' = cor(rna_value, nano_value,
                                    method = 'pearson', use = 'complete'),
                    'spearman' = cor(rna_value, nano_value,
                                     method = 'spearman', use = 'complete'),
                    rna_q50, nano_q50, 'gene_symbol' = sig),
                by = c('timepoint')]
  return(cors)
}), fill = T)

comb_dat <- rbind(mutate(rna_nano_cors, 'type' = 'Single gene'),
                  mutate(rna_nano_cors_sigs, 'type' = 'Gene signature'))
```

```{r, warning=FALSE, echo=FALSE}
plot_dist <- function(dtf = comb_dat[timepoint == 'Baseline'],
                      cor_measure = 'spearman') {
  mycols <- c(maartenutils::darken('lightblue', factor = 1),
              maartenutils::darken('lightblue', factor = 2.5)) %>%
   setNames(dtf[, unique(type)])

  p1 <- ggplot(dtf,
               aes_string(x = cor_measure, y = '(..count..)/sum(..count..)',
                          fill = 'type')) +
    geom_histogram(binwidth = .01, color = 'black', position = 'stack') +
    scale_fill_manual(name = '', values = mycols) +
    xlab(sprintf('%s correlation coefficient', tonic_cap(cor_measure))) +
    ylab('Relative frequency') +
    theme(legend.position = 'top')

  p2 <- ggplot(dtf, aes_string(x = '1', y = cor_measure, fill = 'type')) +
    geom_boxplot(color = 'black') +
    scale_fill_manual(name = '', values = mycols) +
    coord_flip() +
    xlab(' ') +
    ylab(' ') +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          axis.line = element_blank(), legend.position = 'none')

  cowplot::plot_grid(p1, p2, align = 'v', rel_heights = c(7, 1), ncol = 1)
}

plot_dist(cor_measure = 'spearman')
```

```{r, warning=FALSE, echo=FALSE}
plot_dist(cor_measure = 'spearman')
ggsave(filename = file.path(plot_dir, 'spearman_nano_vs_rna_single_genes.pdf'))
plot_dist(cor_measure = 'pearson')
ggsave(filename = file.path(plot_dir, 'pearson_nano_vs_rna_single_genes.pdf'))
```

Plot relationship between correlation and average expression. Are lowly expressed
genes enriched amongst the genes with low concordance between the two
methodologies?


```{r, warning=FALSE, echo=FALSE}
ggplot(unique(comb_dat, by = c('gene_symbol')), 
       aes(x = nano_q50, y = spearman)) + geom_point()
```

```{r, warning=FALSE, echo=FALSE}
library(kableExtra)

unique(comb_dat, by = c('gene_symbol'))[abs(spearman) <= .1] %>%
  .[order(spearman)] %>%
  knitr::kable(format = 'html') %>%
  kableExtra::kable_styling()
```


# CA15.3 correlation with Danaher signatures

```{r, warning=FALSE, echo=FALSE}
## Investigate outlying patient
rna_sigs <- compute_RNA_sigs(sig = 'mmr_loss')
pat_id <- rna_sigs[which.max(FC), patient]
pat_id <- rna_sigs[which.max(FC), patient]
clinical_annotation[patient %in% pat_id]
```

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')

plot_RNA_sig_MA(sig = 'mmr_loss') +
  scale_colour_gradient2(name = 'CA15.3',
                         trans = 'log10',
                         low = scales::muted('red'),
                         high = scales::muted('blue'),
                         na.value = 'white')
```

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
colour_var <- 'ca15_3'

plots <- lapply(danaher_signature_genes[, unique(signature_name)],
                function(sig) {
  plot_RNA_sig_MA(sig = sig, colour_var = colour_var) +
    scale_colour_gradient2(name = 'CA15.3',
                           trans = 'log10',
                           low = scales::muted('red'),
                           high = scales::muted('blue'),
                           na.value = 'white')
})

maartenutils::plot_panel_layout(plots, ncol = 2, nrow = 4,
  labels = NULL,
  plot_direct = F,
  ref_panel_idx = 2 * 8,
  clear_redundant_labels = T,
  clear_redundant_legends = T,
  filename = file.path(plot_dir, sprintf('ma_plot_%s.pdf', colour_var)),
  w = 17.4, h = 25, label_size = 8)
```

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
devtools::load_all(file.path('~/libs', 'GSEAgenesets'))
gene_sets <- filter_gmt(gmt_pat = 'h.all', gene_set_pat = '.')

colour_var <- 'ca15_3'
library(purrr)

plots <- imap(gene_sets, function(g, n) {
  p <- plot_RNA_sig_MA(sig = NULL, genes = g, colour_var = colour_var)
  p + scale_colour_gradient2(name = 'CA15.3',
                             trans = 'log10',
                             low = scales::muted('red'),
                             high = scales::muted('blue'),
                             na.value = 'white') +
    ggtitle(n)
})

maartenutils::plot_panel_layout(plots, ncol = 2, nrow = 4,
  labels = NULL,
  plot_direct = F,
  ref_panel_idx = 2 * 8,
  clear_redundant_labels = T,
  clear_redundant_legends = T,
  filename = file.path(plot_dir, sprintf('ma_plot_hallmark_%s.pdf', colour_var)),
  w = 17.4, h = 25, label_size = 8)
```
