---
title: "Identifying clinical responders from NanoString data in the TONIC trial"
author: "Maarten Slagter"
output: html_document
---

```{r, cache = F, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = F, cache.lazy = F,
                      cache.comments = F, autodep = T, warning = FALSE,
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
```

```{r, warning=FALSE, echo=FALSE, include=F}
source('R/init.R')
```

```{r, eval = F, warning=FALSE, echo=FALSE}
# install.packages('extrafont')
# font_import(pattern="[H/h]elvetica", prompt=FALSE)
# font_import(pattern="helvetica", prompt=FALSE)
# require(ggplot2)
```

# Association with response

RG = genes associated with response on nivo
UG = genes upregulated due to induction treatment, subset of RG

```{r, warning=FALSE, echo=FALSE}
source('R/test_associations.R')
p_vals <- list()

p_vals[['gene_set']] <-
  lapply(auto_name(timepoints), function(tp) {
    res <- data.table(gene_set = test_gene_sets)
    res <- rbindlist(lapply(test_gene_sets, function(gs) {
      test_gene_set_association(gene_set = gs, timepoint = tp,
                                y_var = 'response')
    }))
    return(res[order(-p_val)])
  })

p_vals[['single_gene']] <-
  lapply(auto_name(timepoints), function(tp) {
    res <- data.table(gene_symbol = test_genes)
    res[, 'p_val' := test_gene_association(gene = .SD[, gene_symbol],
                                           timepoint = tp,
                                           y_var = 'response'),
         by = 1:nrow(res)]
    return(res[order(-p_val)])
  })
saveRDS(p_vals, file.path('rds', 'p_vals.rds'))
```

```{r, warning=FALSE, echo=FALSE}
p_vals[['single_gene']][['On nivo']][p_val <= .05]
p_vals[['gene_set']][['Baseline']][p_val <= .05]
p_vals[['gene_set']][['Post-induction']][p_val <= .05]
p_vals[['gene_set']][['On nivo']][p_val <= .05]
p_vals[['single_gene']]$`On nivo`
```

```{r, exporttofiles, warning=FALSE, echo=FALSE}
## Write tables to HTML
pacman::p_load(tableHTML)

dir.create(file.path(p_root, 'html'))
for (s1 in names(p_vals)) {
  for (s2 in names(p_vals[[s1]])) {

    cp_friendly_tp <- timepoints_inv[s2]
    fn <- file.path(p_root, 'html', sprintf('cors_%s_%s.html',
                                            s1, cp_friendly_tp))
    print(fn)
    write_tableHTML(tableHTML(p_vals[[s1]][[s2]]), file = fn)
  }
}
```

## Heatmap of p-values

```{r, warning=FALSE, echo=FALSE}
m <- rbindlist(lapply(auto_name(sig_gene_sets), function(gs) {
  test_gene_set_association
}), fill = T)
m[p_val > .05, p_val := NA]

p <- plot_p_values_induction(m)
ggsave(plot = p, filename = file.path(plot_dir,
                            'induction_fc_p_values_baseline_on_nivo.png'),
       width = 3.3, height = 5.6, units = 'in')
```

# Overlap genes in gene signatures

Virtually no overlap in genes involved in gene signatures

```{r, warning=FALSE, echo=FALSE}
m <- dcast(danaher_signature_genes, signature_name ~ gene_name,
           fun.aggregate = length) %>%
  as.data.frame %>%
  { column_to_rownames(., var = 'signature_name') } %>%
  as.matrix %>%
  t %>%
  overlap_analysis
plot(m)
```

Virtually no overlap in genes involved in gene signatures

```{r, warning=FALSE, echo=FALSE}
devtools::load_all(file.path('~/libs', 'maartenutils'))

danaher_signature_genes[, 'signature_name' := gsub(' ', '.',
                                                   tolower(signature_name))]
danaher_signature_genes[, 'signature_name' := gsub('\\.|-', '.',
                                                   tolower(signature_name))]
stopifnot(all(sig_gene_sets %in% 
              danaher_signature_genes[, unique(signature_name)]))

m <- dcast(danaher_signature_genes[signature_name %in% sig_gene_sets],
           formula = signature_name ~ gene_name,
           fun.aggregate = length) %>%
  as.data.frame %>%
  { column_to_rownames(., var = 'signature_name') } %>%
  as.matrix %>%
  t %>%
  overlap_analysis

p <- plot(m, cap_fun = identity)
```

# Parallel coordinates plots

## Gene level

```{r, warning=FALSE, echo=FALSE}
source('R/load_data.R')
source('R/plotting.R')
colour_var = 'response'
exp_levels[, gene_symbol := gsub('-|\\.', '-', gene_symbol)]
genes = exp_levels[, sort(unique(gene_symbol))] %>%
  # qs %>%
  as.character

N_plots <- length(genes)
ppp <- 16
N_pages <- ceiling(N_plots / ppp)
N_brackets <- (ceiling(N_plots / ppp) * ppp)
plot_idx <- suppressWarnings(base::split(1:N_plots, rep(1:N_pages, each = ppp)))

plyr::l_ply(seq_along(plot_idx), function(idx) {
  li <- plot_idx[[idx]]
  if (length(li) == 0 || all(is.na(li))) return(NULL)
  li_first <- li[1]
  plots <- lapply(li, function(x) {
    p <- plot_parallel_coords_single_gene(genes[x], colour_var = colour_var) +
      ggplot2::theme(legend.position = 'none') +
      scale_y_continuous(name = 'log2 expression')
    if (x == li_first) {
      p <- p + theme(legend.direction = 'vertical', legend.position = 'right')
    }
    return(p)
  })
  fn <- sprintf('plots/parallel_single_gene_idx-%s_%s.pdf', idx, colour_var)
  plot_panel_layout(plots, ncol = 4, nrow = 4,
                    labels = NULL,
                    filename = fn,
                    w = 22, h = length(plots) / 5 * 11,
                    label_size = 8)
})
```

## Geneset level

Which induction arms most strongly upregulate gene signatures associated with
response?

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
colour_var = NULL
colour_var = 'patient'
colour_var = 'response'

plots <- lapply(danaher_scores.m[, unique(variable)], function(x) {
  p <- plot_parallel_coords_geneset(x, colour_var = colour_var,
                                    facet_var = 'arm')
  return(p)
})
```

```{r, warning=FALSE, echo=FALSE}
filename <- file.path(img_dir, sprintf('parallel_%s.pdf',
                                       ifelse(is.null(colour_var), 'ungrouped',
                                              colour_var)))

maartenutils::plot_panel_layout(plots, ncol = 5, filename = filename,
                                labels = NULL,
                                w = 22, h = length(plots) / 5 * 9,
                                label_size = 8)
sys_file_open(filename)
```

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_nanostring.R')

gen_parallel_coord_plots <- 
  function(facet_var = 'arm', colour_var = 'patient') {
  plots <- lapply(danaher_scores.m[, auto_name(levels(variable))],
                  function(gene_set) {
                    browser()
    sum_dat <- prep_geneset_parallel_coords(gene_set = gene_set,
                                            colour_var = facet_var,
                                            facet_var = facet_var)
    p_dat <- danaher_scores.m[variable == gene_set]
    p <- plot_parallel_coords(p_dat = p_dat, sum_dat = sum_dat,
                              palette_name = 'FantasticFox',
                              facet_var = facet_var,
                              point_alpha = .6,
                              colour_var = colour_var, title = gene_set) +
      theme(legend.position = 'none')
    return(p)
  })

  filename <- file.path(img_dir, sprintf('parallel_%s.pdf',
                                         ifelse(is.null(colour_var), 'ungrouped',
                                                colour_var)))
  p <- maartenutils::plot_panel_layout(plots, ncol = 1, nrow = 4,
                                       filename = filename,
                                       labels = NULL,
                                       w = 22, h = 25)

  filename <- file.path(img_dir, sprintf('parallel_%s_highlighted.pdf',
                                         ifelse(is.null(colour_var), 'ungrouped',
                                                colour_var)))
  p <- maartenutils::plot_panel_layout(
    list(plots[['tis']], plots[['glycolytic.activity']]),
    ncol = 1, nrow = 2,
    filename = filename,
    labels = NULL,
    w = 22, h = 25 / 2)
  return(plots)
}

# plots <- gen_parallel_coord_plots(facet_var = 'arm', colour_var = 'patient')
plots <- gen_parallel_coord_plots(facet_var = 'arm', colour_var = 'response')
```

```{r, warning=FALSE, echo=FALSE}
```

```{r, warning=FALSE, echo=FALSE}
devtools::load_all(file.path('~/libs', 'maartenutils'))
library(grid)
plot_panel_layout(lapply(sig_gene_sets, function(gs) {
                           p <- plot_parallel_coords_geneset(gene_set = gs)
                           if (gs != sig_gene_sets[1])
                             p <- p + ggplot2::theme(legend.position = 'none')
                           else
                             p <- p + ggplot2::theme(legend.position = 'right')
                           return(p)
                         }),
                  filename = file.path('plots', 'parallel_sig_signatures.pdf'),
                  labels = NULL,
                  w = 17.4, h = 25, ncol = 3)
```

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
devtools::load_all('~/libs/maartenutils')
plot_parallel_coords_geneset(gene_set = 'tis', facet_var = 'arm')
```

Compare tp 2 to tp 0 in terms of expression of significant gene sets

Raw data in combination with association with response

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')

plots <- lapply(auto_name(test_gene_sets), function(gs) {
  offs = grid::unit(.05, 'npc')
  label_size = 8
  if (gs %in% sig_gene_sets) {
    resp_ass <- grid::textGrob('R',
                               x = 2 * offs,
                               y = grid::unit(1, 'npc') - offs,
                               gp = grid::gpar(fontsize = label_size,
                                               col='black',
                                               fontface = 'bold'))
  } else {
    resp_ass <- NULL
  }
  # resp_ass_sq <- grid::rectGrob('A', x = offs,
  #                            y = grid::unit(1, 'npc') - offs,
  #                            gp = grid::gpar(fontsize = label_size,
  #                                            col='black',
  #                                            fontface = 'bold'))
  main_p <- plot_parallel_coords_geneset(gene_set = gs, facet_var = 'arm')
  if (!gs == test_gene_sets[1])
    main_p <- main_p + theme('legend.position' = 'none')
  return(grid::grobTree(to_g(main_p), resp_ass))
})

ml <- gridExtra::marrangeGrob(plots, nrow=4, ncol=1, top = '')
ggsave(plot = ml,
       filename = file.path(plot_dir, 'parallel_genesets_by_arm.pdf'),
       width = 17.4, height = 25, units = 'cm')
```

Heatmaps of p-values

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
m <- rbindlist(lapply(auto_name(sig_gene_sets), function(gs) {
  test_gene_set_difference(gene_set = gs, tp1 = 'Baseline', tp = 'Post-induction')
}), fill = T)
invisible(m[p_val > .05, p_val := NA])
p <- plot_p_values_induction(m)
ggsave(plot = p, filename = file.path(plot_dir,
                            'induction_fc_p_values_baseline_post_induction.png'),
       width = 10, height = 13, units = 'cm')
```

```{r, warning=FALSE, echo=FALSE}
m <- rbindlist(lapply(auto_name(sig_gene_sets), function(gs) {
  test_gene_set_difference(gene_set = gs, tp1 = 'Baseline', tp = 'On nivo')
}), fill = T)
invisible(m[p_val > .05, p_val := NA])

p <- plot_p_values_induction(m)
ggsave(plot = p, filename = file.path(plot_dir,
                            'induction_fc_p_values_baseline_on_nivo.png'),
       width = 10, height = 13, units = 'cm')
```

# Correlations in gene expression set scores

```{r, warning=FALSE, echo=FALSE}
library(corrplot)

cor_mat <-
  danaher_scores.m[, dcast(data = .SD, formula = patient ~ variable,
                           value.var = 'value'), by = timepoint]

plot_cor_mat <- function(tp = 'Baseline',
                         cormethod = 'spearman') {
  cor_mat[timepoint == tp] %>%
    dplyr::select(-timepoint, -patient) %>%
    { .[apply(., 1, function(x) all(!is.na(x)))] } %>%
    as.matrix %>% cor(., method = cormethod) %>%
    corrplot(., order = "hclust", addrect = 6, tl.col = 'gray40',
             tl.pos = 'ld', type = 'lower',
             tl.cex = .8,
             # addCoef.col = "black", # Add coefficient of correlation
                      # Combine with significance
             # sig.level = 0.05, insig = "blank",
             mar=c(0,0,1,0), # http://stackoverflow.com/a/14754408/54964
             title = sprintf('%s - %s', tp, simple_cap(cormethod)))
}

pdf(file = file.path(plot_dir, 'danaher_signature_correlations.pdf'),
    paper = 'a4', width = 0, height = 0)
layout(matrix(1:6, byrow = T, nrow = 3))
plot_cor_mat(tp = 'Baseline', cormethod = 'pearson')
plot_cor_mat(tp = 'Baseline', cormethod = 'spearman')
plot_cor_mat(tp = 'Post-induction', cormethod = 'pearson')
plot_cor_mat(tp = 'Post-induction', cormethod = 'spearman')
plot_cor_mat(tp = 'On nivo', cormethod = 'pearson')
plot_cor_mat(tp = 'On nivo', cormethod = 'spearman')
dev.off()
```

# Test pat 22

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
patient_labels[patient == 'pat_26']
prepare_test_gene_set_difference(gene_set = 'mmr.loss',
                                 facet_var = NULL,
                                 tp1 = 'Baseline',
                                 tp2 = 'On nivo')
prepare_test_gene_set_difference(gene_set = 'mmr.loss',
                                 facet_var = NULL,
                                 tp1 = 'Baseline',
                                 tp2 = 'Post-induction')
```
