---
title: "Identifying clinical responders from RNASeq data in the TONIC trial"
author: "Maarten Slagter"
output: html_document
---

```{r, cache = F, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = F, cache.lazy = F,
                      cache.comments = F, autodep = T, warning = FALSE,
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
```

```{r, warning=FALSE, echo=FALSE, include=F}
source('R/init.R')
source('R/load_rna_dat.R')
source('R/ML_RNASeq.R')
```

# Check quality of data

## PCA approach

```{r, warning=FALSE, echo=FALSE}
# rownames(rna_read_counts_salmon)
# dimnames(rna_read_counts_salmon)[[1]] <- rownames(rna_read_counts_salmon)
run_pca <- function(dtf = rna_read_counts_salmon) {
  t_dtf <- as.matrix(dtf) %>% t 
  colnames(t_dtf) <- rownames(dtf)
  t_dtf <- t_dtf %>%  { .[, !eps(apply(., 2, var), 0, 1e-1)] }
  # dimnames(t_rna_read_counts_salmon)
  pca <- prcomp(t_dtf, center = T, scale = T)
  return(pca)
}
pca <- run_pca(dtf = tpms_salmon)
plot(pca, type = "l")
```

```{r, warning=FALSE, echo=FALSE}
library(ggbiplot)

g <- ggbiplot(pca, obs.scale = 1, var.scale = 1, 
              # groups = rna_sample_annotation[match(rownames(t_rna_read_counts_salmon), cf_number), timepoint], 
              groups =
                rna_sample_annotation[match(rownames(t_rna_read_counts_salmon),
                                            cf_number), arm], 
              # groups =
              #   rna_sample_annotation[match(rownames(t_rna_read_counts_salmon),
              #                               cf_number), response], 
              var.axes = F, 
              arrow = F,
              ellipse = F, circle = F)
# g <- g + scale_color_discrete(name = '')
# g <- g + theme(legend.direction = 'horizontal', legend.position = 'top')
print(g)
```

## Other approaches

```{r, warning=FALSE, echo=FALSE}
apply(tpms_salmon, 2, compute_evenness)
apply(tpms_salmon, 2, median)
```

# Glmnet approach

```{r, warning=FALSE, echo=FALSE}
glmnet_class_test_rnaseq(tp = 'Baseline', type = 'genes')
```

```{r, warning=FALSE, echo=FALSE}
glmnet_class_test_rnaseq(tp = 'Post-induction', type = 'genes')
```

```{r, warning=FALSE, echo=FALSE}
layout(matrix(1:2, ncol = 2, nrow = 1, byrow = T))
r_b <- glmnet_class_test_rnaseq(tp = 'Baseline', type = 'genes', alphas = 1)
r_pi <- glmnet_class_test_rnaseq(tp = 'Post-induction', type = 'genes', alphas = 1)
```

```{r, warning=FALSE, echo=FALSE}
plot(r_b$glmnet.fit)
r_b %>% 
  { coef(.$glmnet.fit, .$lambda.min) } %>%
  { setNames(.[which(. != 0)], rownames(.)[which(. != 0)]) }

r_pi %>% 
  { coef(.$glmnet.fit, .$lambda.min) } %>%
  { setNames(.[which(. != 0)], rownames(.)[which(. != 0)]) } %>%
  { .[order(abs(.))] }
```

# SVM approach to training

```{r, warning=FALSE, echo=FALSE}
source('R/ML_RNASeq.R')
t_dat <- prepare_ml_dat_rnaseq(tp = 'Post-induction', type = 'genes')
```

```{r, warning=FALSE, echo=FALSE}
ctrl <- trainControl(method = 'repeatedcv',   # 10fold cross validation
                     repeats = 1,				      # do 5 repititions of cv
                     # Use AUC to pick the best model
                     summaryFunction = twoClassSummary,	
                     classProbs = TRUE)

svm.tune <- train(x = t_dat$t_dat,
                  y = t_dat$responses,
                  method = 'svmRadial',           # Radial kernel
                  tuneLength = 9,					        # 9 values of the cost function
                  preProc = c('center','scale'),  # Center and scale data
                  metric = 'ROC',
                  trControl = ctrl)
```

```{r, warning=FALSE, echo=FALSE}
plot(svm.tune)
confusionMatrix(svm.tune)
```

# Naive/reference model performance

Naive model classification performance
Classify according to response rate

```{r, warning=FALSE, echo=FALSE}
p <- mean(as.logical(responses))
p;^2 + (1 - p)^2
2 * p * (1 - p)
curve(expr = x^2 + (1 - x)^2, from = 0, to = 1)
# curve(expr = 2 * x * (1 - x), from = 0, to = 1)
```

# PCA approach

Response seems to be partially explained by the first two principal components
on therapy samples.
For timepoints later than 1, we should probably be looking at fold changes
instead.

```{r, warning=FALSE, echo=FALSE}
ml_dat <- prepare_ml_dat(tp = 'On nivo')
plot_PCA_axes(tp = 'on_nivo', dat = ml_dat[[1]], color_var = 'response',
              labellings = ml_dat[[2]])

ml_dat <- prepare_ml_dat(tp = 'Post-induction')
plot_PCA_axes(tp = 'post_induction', dat = ml_dat[[1]], color_var = 'response',
              labellings = ml_dat[[2]])

ml_dat <- prepare_ml_dat(tp = 'Baseline')
plot_PCA_axes(tp = 'baseline', dat = ml_dat[[1]], color_var = 'response',
              labellings = ml_dat[[2]])

setkey(patient_labels, filename)
ml_dat <- prepare_ml_dat(tp = 'Post-induction')
plot_PCA_axes(tp = 'post_induction', dat = ml_dat[[1]], color_var = 'arm',
              labellings = patient_labels[rownames(ml_dat[[1]]), arm])

setkey(patient_labels, filename)
ml_dat <- prepare_ml_dat(tp = 'Baseline')
plot_PCA_axes(tp = 'baseline', dat = ml_dat[[1]], color_var = 'arm',
              labellings = patient_labels[rownames(ml_dat[[1]]), arm])
```

# Random forest approach

Given that gene sets associated with clinical response at timepoint 2 were also
already higher at baseline (when combining all treatment arms), it seems
plausible that responders can be identified at baseline based on these gene
sets.

```{r, cache = T, dependson, warning=FALSE, echo=FALSE}
source('R/ML.R')
rfs <- plyr::llply(timepoints, function(tp) {
  perform_rf(tp = tp)
}, .parallel = !local_run)
```

```{r, warning=FALSE, echo=FALSE}
plots <- lapply(rfs, plot_importance)
plot_panel_layout(plots, ncol = 3, filename = NULL,
                  w = 20, h = 20, label_size = 8)
```

Use clinical information to train predictor

```{r, warning=FALSE, echo=FALSE}
source('R/init.R')
library(maartenutils)

x_dat <- clinical_annotation[, .(induction_therapy, who_performance_status,
                                 s_til, ca15_3, response_bin)] %>%
  maartenutils::filter_na_rows()
invisible(x_dat[, response_bin := factor(response_bin)])

rf_default <- caret::train(x = as.data.frame(select(x_dat, -response_bin)),
  y = unlist(select(x_dat, response_bin)),
  method = "rf",
  tuneGrid = expand.grid(.mtry=ceiling(sqrt(ncol(x_dat - 1)))),
  trControl = trainControl(method="repeatedcv",
  number=nrow(x_dat),
  savePredictions = T,
  repeats=10))
```

Kappa of zero indicates no added benefit over a 'null' classifier which simply
classifies randomly based on incidence.

```{r, warning=FALSE, echo=FALSE}
str(rf_default)
selectedIndices <- rf_default$pred$mtry == 3
str(rf_default)
rf_default$pred
library(caret)
library(pROC)
head(rf_default$pred)

plot.roc(rf_default$pred$obs[selectedIndices],
         rf_default$pred$M)

```
