---
title: "TIL and blood TCR repertoires and their relationships to induction arm and clinical response"
author: "Maarten Slagter"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: united
---


```{r, warning=FALSE, echo=FALSE, cache = F}
knitr::opts_chunk$set(message = FALSE, cache = T, cache.lazy = F,
                      cache.comments = F, autodep = T, warning = FALSE,
                      results = 'hide', fig.keep = 'high',
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# knitr::clean_cache()
# knitr::opts_chunk$get("cache.path")
# knitr::knit_global()
source('R/init.R')
```

# Relationship between various repertoire size estimators and platforms

Various (statistical) methods of estimating the total repertoire size,
extrapolating from the observed to the unobserved total.

```{r, fig.width = 12, fig.height = 6, warning=FALSE, echo=FALSE}
plot_cor_mat()
## Separating by timepoint does not change these correlations for the Adapative
## data, as expected
plot_cor_mat(tp = 'Baseline')
# plot_cor_mat(tp = 'Post-induction')
# plot_cor_mat(tp = 'On nivo')
```

# Single timepoint summary statistic associations

Highly interesting to me: repertoire size is correlated with response at
baseline, but not post-induction or on nivo.
On the contrary, clonality is associated with response post-induction and on
nivo, T-cell infiltration only on-nivo.

## Tumor

```{r, warning=FALSE, echo=FALSE, results='asis'}
res <- test_adaptive_association(timepoint = 'Baseline')
print(knitr::kable(res))
res <- test_adaptive_association(timepoint = 'Post-induction')
print(knitr::kable(res))
res <- test_adaptive_association(timepoint = 'On nivo')
print(knitr::kable(res))
```

Visualisations of the comparisons shown above in table form

```{r, warning=FALSE, echo=FALSE, fig.width = 3, fig.height = 4}
plot_single_timepoint_r_assocation(l_timepoint = 'Baseline', 
                                   facet_var = NULL)
plot_single_timepoint_r_assocation(l_timepoint = 'Post-induction', 
                                   facet_var = NULL)
plot_single_timepoint_r_assocation(l_timepoint = 'On nivo', 
                                   facet_var = NULL)
```

```{r, warning=FALSE, echo=FALSE}
```

The same comparisons but split up by induction arm

```{r, warning=FALSE, echo=FALSE, fig.width = 5, fig.height = 4}
plot_single_timepoint_r_assocation(l_timepoint = 'Baseline', 
                                   facet_var = 'arm')
plot_single_timepoint_r_assocation(l_timepoint = 'Post-induction', 
                                   facet_var = 'arm')
plot_single_timepoint_r_assocation(l_timepoint = 'On nivo', 
                                   facet_var = 'arm')
```

## Blood

```{r, warning=FALSE, echo=FALSE, results='asis'}
res <- test_adaptive_association(timepoint = '-2')
print(knitr::kable(res))
res <- test_adaptive_association(timepoint = '0')
print(knitr::kable(res))
res <- test_adaptive_association(timepoint = '6')
print(knitr::kable(res))
# res <- test_adaptive_association(timepoint = '12')
# print(knitr::kable(res))
```


```{r, warning=FALSE, echo=FALSE, fig.width = 3, fig.height = 4}
plot_single_timepoint_r_assocation(l_timepoint = '-2', 
                                   facet_var = NULL)
plot_single_timepoint_r_assocation(l_timepoint = '0', 
                                   facet_var = NULL)
plot_single_timepoint_r_assocation(l_timepoint = '6', 
                                   facet_var = NULL)
```

```{r, warning=FALSE, echo=FALSE, fig.width = 5, fig.height = 4}
plot_single_timepoint_r_assocation(l_timepoint = '-2', 
                                   facet_var = 'arm')
plot_single_timepoint_r_assocation(l_timepoint = '0', 
                                   facet_var = 'arm')
plot_single_timepoint_r_assocation(l_timepoint = '6', 
                                   facet_var = 'arm')
```

# Relationship between distribution and induction treatment & clinical response

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_adaptive.R')
source('R/plotting_nanostring.R')
plot_adaptive_FC(facet_var = 'arm', 
                 p_timepoints = c('Baseline', 'Post-induction', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'comb_time_resp')
```

```{r, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm', 
                 p_timepoints = c('Baseline', 'Post-induction'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'comb_time_resp')
```

```{r, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm', 
                 p_timepoints = c('Baseline', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'comb_time_resp')
```

The same, but not coloring by clinical response

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_adaptive.R')
plot_adaptive_FC(facet_var = 'arm', 
                 p_timepoints = c('Baseline', 'Post-induction', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'timepoint')
```

```{r, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm', 
                 p_timepoints = c('Baseline', 'Post-induction'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'timepoint')
```

```{r, warning=FALSE, echo=FALSE}
plot_adaptive_FC(facet_var = 'arm', 
                 p_timepoints = c('Baseline', 'On nivo'),
                 var1 = 'sample_clonality', var2 = 'efron_thisted_estimator',
                 colour_var = 'timepoint')
```

```{r, eval = F, warning=FALSE, echo=FALSE}
s_plot <- ggplot(patient_labels, aes(x = efron_thisted_estimator,
                                     y = sample_clonality,
                                     label = timepoint_number,
                                     shape = comb_time_resp,
                                     group = patient,
                                     colour = comb_time_resp)) +
  geom_line(alpha = alpha_lev, show.legend = F) +
  geom_point(size = 2, alpha = alpha_lev) +
  # ggrepel::geom_text_repel(segment.color = 'grey90', color = 'black') +
  # geom_text(color = 'white') +
  # geom_smooth(se = F) +
  scale_fill_manual(values = comb_time_resp_palette) +
  scale_colour_manual(name = '', values = comb_time_resp_palette) +
  scale_shape_manual(name = '',
                     values = setNames(c(1, 2, 3, 1, 2, 3) + 14,
                                       names(comb_time_resp_palette))) +
  scale_x_continuous('Repertoire size', trans = 'log10', expand = c(0, 0)) +
  scale_y_continuous('Repertoire evenness', trans = 'identity',
                     expand = c(0, 0)) +
  theme(legend.position = 'top', legend.direction = 'vertical') +
  gg_legend_alpha_cancel
  # guides(shape = guide_legend(nrow = 3))

if (!is.null(facet_var)) {
  s_plot <- s_plot + facet_wrap(as.formula(sprintf('~ %s',
                                                   paste(facet_var,
                                                         collapse = ' + '))),
                                nrow = length(facet_var))
}
print(s_plot)
```

```{r, warning=FALSE, echo=FALSE, fig.width = 4, fig.height = 6}
source('R/plotting_nanostring.R')
source('R/plotting_adaptive.R')
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'tumor',
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'tumor', 
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'tumor', 
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'blood',
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'blood',
                       facet_var = NULL)
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'blood',
                       facet_var = NULL)
```

The same but restricting to some time points for clarity: -2 (pre) and 6 (on
nivo)

```{r, warning=FALSE, echo=FALSE, fig.width = 4, fig.height = 6}
source('R/plotting_nanostring.R')
source('R/plotting_adaptive.R')
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'blood',
                       facet_var = NULL, 
                       allowed_timepoints = c(-2, 6))
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'blood',
                       facet_var = NULL, 
                       allowed_timepoints = c(-2, 6))
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'blood',
                       facet_var = NULL, 
                       allowed_timepoints = c(-2, 6))
```



```{r, fig.width = 7, fig.height = 3, warning=FALSE, echo=FALSE}
source('R/plotting_nanostring.R')
source('R/plotting_adaptive.R')
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'tumor',
                       facet_var = 'arm')
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'tumor', 
                       facet_var = 'arm')
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'tumor', 
                       facet_var = 'arm')
```

```{r, fig.width = 7, fig.height = 3, warning=FALSE, echo=FALSE}
# debugonce(plot_parallel_adaptive)
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', 
                       compartment = 'blood',
                       facet_var = 'arm')
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'blood',
                       facet_var = 'arm')
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'blood',
                       facet_var = 'arm')
```

```{r, fig.width = 7, fig.height = 3, warning=FALSE, echo=FALSE}
plot_parallel_adaptive(p_var = 'efron_thisted_estimator', compartment = 'blood',
                       facet_var = 'arm',
                       allowed_timepoints = c(-2, 6))
plot_parallel_adaptive(p_var = 'sample_clonality', compartment = 'blood',
                       facet_var = 'arm',
                       allowed_timepoints = c(-2, 6))
plot_parallel_adaptive(p_var = 'adaptive_t_cells', compartment = 'blood',
                       facet_var = 'arm',
                       allowed_timepoints = c(-2, 6))
```


```{r, eval = F, warning=FALSE, echo=FALSE}
source('R/plotting_nanostring.R')
cowplot::plot_grid(
  plot_parallel_adaptive(p_var = 'efron_thisted_estimator') + remove_x +
    remove_legend + scale_y_continuous(trans = 'log10',
                                       name = 'Repertoire size'),
  plot_parallel_adaptive(p_var = 'sample_clonality') + remove_strip +
    theme(legend.position = 'top'),
  s_plot + remove_strip, ncol = 1, align = 'hv', rel_heights = c(.4, .4, .6))
```

# Deltas in summary statistics over time

```{r, warning=FALSE, echo=FALSE}
plot_all_comps(f = compare_adaptive_summary_stats,
               colour_var = 'clinical_response',
               x_var = 'clinical_response',
               facet_var = NULL)
```

Effect of induction arm on these measures

```{r, eval = T, warning=FALSE, echo=FALSE}
# graphics.off()
source('R/plotting_nanostring.R')
source('R/plotting_adaptive.R')
plot_all_comps(f = compare_adaptive_summary_stats,
               colour_var = 'arm',
               x_var = 'arm',
               facet_var = NULL)
```

Split out by clinical response, contrasts most pronounced in responders

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_nanostring.R')
source('R/plotting_adaptive.R')
plot_all_comps(f = compare_adaptive_summary_stats,
               colour_var = 'clinical_response',
               x_var = 'clinical_response',
               facet_var = 'arm')
```

# Clone size FCs compared between tumor and blood

```{r, warning=FALSE, echo=FALSE, results = 'hide'}
plot_tp_comp_FCs(tp1 = 'Baseline', tp2 = 'Post-induction', 
                 tp3 = '-2', tp4 = '0')
plot_tp_comp_FCs(tp1 = 'Baseline', tp2 = 'On nivo', 
                 tp3 = '-2', tp4 = '6')
plot_tp_comp_FCs(tp1 = 'Baseline', tp2 = 'On nivo', 
                 tp3 = '-2', tp4 = '10')
plot_tp_comp_FCs(tp1 = 'Baseline', tp2 = 'On nivo', 
                 tp3 = '-2', tp4 = '12')
```

# Direct comparison of clone sizes in tumor and blood

```{r, warning=FALSE, echo=FALSE, results = 'hide'}
# devtools::load_all(file.path('~/libs', 'maartenutils'))
source('R/plotting_nanostring.R')
source('R/plotting_adaptive.R')
plot_tp_comp_direct(tp1 = 'Baseline', tp2 = '-2')
plot_tp_comp_direct(tp1 = 'Post-induction', tp2 = '0')
plot_tp_comp_direct(tp1 = 'On nivo', tp2 = '6')
plot_tp_comp_direct(tp1 = 'On nivo', tp2 = '10')
plot_tp_comp_direct(tp1 = 'On nivo', tp2 = '12')
```

# Tracing individual TCRs

We should be interested in clone size relative to all (nucleated) cells, rather
than relative to the other T cells, but the latter is the default metric that
Adaptive reports (i.e. productive frequency). Productive frequency is however
problematic when comparing samples from the same patient - as we're doing here -
as the relative size of a clone depends on the sizes of the other clones and as
such expansion of a clone will automatically mean the lowering of the relative
contribution of the other clones. I corrected the relative contribution by
multiplying it with the fraction of T-cells over all nucleated cells, which "is
calculated by dividing the number of Productive Templates by Nucleated Cells (a
measure based on the amplification of reference gene primers rather than DNA
absorbance)." 

```{r, warning=FALSE, echo=FALSE}
plot_TCR_chronological(patient = 'pat_11', 
                       p_var = 'productive_frequency') + 
  ggtitle('Not normalised for T cell infiltration')
plot_TCR_chronological(patient = 'pat_11',
                       p_var = 'normalized_frequency') +
  ggtitle('Normalised for T cell infiltration')
```

Plotting these kinds of plots for all patients. It seems as if responding
patients consistently have a high amount of high magnitude new clones 'On nivo',
more so than the pesistence and expansion of already existing clones, which also
occurs in non-responding patients.

```{r, tcr_chrono, warning=FALSE, echo=FALSE, fig.height = 4, fig.width = 4}
#' Plot TCR chronological plots for all patients
#'
#'
plot_TCR_chronological_all <- function(p_var = 'normalized_frequency') {
  read_adaptive_seqs(force_reload = F)
  plyr::llply(arr[, naturalsort::naturalsort(auto_name(unique(patient)))], 
              function(patient) {
    p_title <- format_patient_title(patient)
    print(p_title)
    p <- plot_TCR_chronological(patient = patient, p_var = p_var)
    if (is.null(p)) return(NULL)
    else return(p + ggtitle(p_title))
  })
}
plot_TCR_chronological_all()
```

```{r, warning=FALSE, echo=FALSE}
source('R/plotting_nanostring.R')
source('R/plotting_adaptive.R')
analyze_top_clones <- function(p_var = 'normalized_frequency') {
  read_adaptive_seqs(force_reload = F)
  dtf <- rbindlist(
    plyr::llply(arr[, naturalsort::naturalsort(auto_name(unique(patient)))], 
              function(patient) {
    p_title <- format_patient_title(patient)
    print(p_title)
    pat_arr <- prepare_TCR_chrono(patient = patient, 
                                  p_var = p_var)
    if (null_dat(pat_arr)) return(NULL)
    pat_arr <- pat_arr[timepoint == 'On nivo']
    if (null_dat(pat_arr)) return(NULL)
    percentage_new_among_50th <- 
      pat_arr %>%
        { .[timepoint == 'On nivo'] } %>%
        { .[order(value)] } %>%
        { .[(cumsum(clone_N * value) / 
            sum(clone_N * value, na.rm =T)) >= .5] } %>%
        { .[, mean(shared_timepoints == 1)] }
    if (is.nan(percentage_new_among_50th)) browser()
    print(percentage_new_among_50th)
    return(list('patient' = patient, 'frac_new' = percentage_new_among_50th))
  }), fill = T)
  dtf <- controlled_merge(dtf, patient_labels[timepoint == 'On nivo'])
  return(dtf)
}
# clear_object(frac_new_clones, sr = sr)
sr(frac_new_clones <- analyze_top_clones())
```

## Composition of largest clones highly associated with clinical response

Indeed, ordering the clones by size and computing the fraction of newly detected
clones among the clones making up the top 50% of the TIL (roughly the top half
of the plots shown above), we see much higher fractions on nivolumab in
responders than in non-responders.

```{r, warning=FALSE, echo=FALSE, fig.width = 3, fig.height = 3}
facet_var = NULL
new_var_name <- 'Fraction of T cell clones newly detected\namong top 50%'
# frac_new_clones[, as.integer(clinical_response)]
frac_new_clones[, 'x_coord' := vipor::offsetX(frac_new, width = .4) + as.integer(clinical_response)] 

p <- ggplot(frac_new_clones,
       aes_string(x = 'x_coord', y = 'frac_new',
                  fill = 'clinical_response')) +
  geom_boxplot(outlier.size = 0) +
  ggbeeswarm::geom_quasirandom() +
  ggpubr::stat_compare_means(label = 'p.signif') +
  scale_fill_manual(name = '', values = resp_colors, guide = F) +
  scale_y_continuous(name = new_var_name) +
  scale_x_continuous(name = '', breaks = c(1, 2), 
                     labels = frac_new_clones[, levels(clinical_response)]) +
  xlab('Clinical response') + 
  
  ggrepel::geom_label_repel(data = frac_new_clones[clinical_response == 'NR' &
                            frac_new >= frac_new_clones[clinical_response == 'R', min(frac_new)]],
             aes(label = patient), fill = 'white', size = 2)
print(p)
```

```{r, warning=FALSE, echo=FALSE, fig.width = 5, fig.height = 3}
p + facet_grid(as.formula(sprintf('~ %s ', 'arm')))
frac_new_clones[patient == 'pat_48']
frac_new_clones[frac_new > .59]
```

This variable adds a little bit of information over the sample clonality
measure, as the two are only weakly correlated. 

```{r, warning=FALSE, echo=FALSE, fig.width = 3, fig.height = 3}
ggplot(frac_new_clones, aes(x = sample_clonality, y = frac_new, 
                            colour = clinical_response)) +
  geom_point() +
  theme(aspect.ratio = 1) +
  scale_colour_manual(name = '',
                      values = tonic_color_palettes[['clinical_response']]) +
  xlab('Sample clonality') + 
  ylab(new_var_name)
```
