---
title: "Permutation testing for enrichment of upregulation of immune associated genes in the Dox and Cis treatment arms"
author: "Maarten Slagter"
output: html_document
---

```{r, cache = F, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = F, cache.lazy = F, 
                      cache.comments = F, autodep = T, warning = FALSE, 
                      echo = TRUE, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
```

Ton was especially interested in the pattern of all gene sets associated with
clinical response moving up and down jointly (see logFC heatmaps). To test the
enrichment with which this occurs in the two arms that came out as winners in
this visualisation (Dox and Cis), we devised a permutation experiment in which
patients are shuffled across treatment arms. First, to summarise the pattern of
gene set fold changes for each patient individually, the median $FC_p$ of fold
changes between two time points of the 16 gene sets was computed to capture the
general change in gene expression in these gene sets. Next, the treatment arms
were summarized by taking the median of $FC_p$ for the patients in that
treatment arm to get $FC_t$. Only patients with samples at both timepoints under
comparison are included in the analysis. To obtain a null distribution of
$FC_t$, $FC_{t\ast}$, that would have occurred given no differential effect of
the treatment arms, arm labels were repetitively shuffled between patients ($n =
1000$). $FC_t$ could then be compared against $FC_{t\ast}$, taking the arm
specific percentage of $FC_{t\ast} \ge FC_t$ as the one-sided probability that
the observed enrichment of fold changes - or lack thereof - could occur by
chance. We could also turn this into a two sided test if desired but to me this
is most intuitive.

Colored in dark red are arm specific median FC in permuted samples that are
greater than the observed FC. Interestingly, the strong effect observed when
comparing $t = 0$ to $t = 2$ for the Dox and Cisplating arms can already be
observed between $t = 0$ and $t = 1$, albeit to a smaller degree. 

A caveat in the analysis of these results is that this method does not take
differences in baseline features into account.

# Results

Baseline vs. post-induction

```{r, warning=FALSE, echo=FALSE, fig.width = 10 / 2.54, height = 13 / 2.54}
source('R/init.R')
library(grid)

p <- ton_percentile_comparison_of_fc(N_reps = 5e4, 
                                     gene_sets = sig_gene_sets,
                                     tp1 = 'Baseline', 
                                     tp2 = 'Post-induction')
grid::grid.draw(p)
```

Post-induction vs. on nivo


```{r, warning=FALSE, echo=FALSE, fig.width = 10 / 2.54, height = 13 / 2.54}
p <- ton_percentile_comparison_of_fc(N_reps = 5e4, 
                                     tp1 = 'Post-induction', 
                                     tp2 = 'On nivo')
grid::grid.draw(p)
```

Baseline vs. on nivo

```{r, warning=FALSE, echo=FALSE, fig.width = 10 / 2.54, height = 13 / 2.54}
p <- ton_percentile_comparison_of_fc(N_reps = 5e4, 
                                     tp1 = 'Baseline', 
                                     tp2 = 'On nivo')
grid::grid.draw(p)
```
