---
title: "Potential confounding variables in transcriptomic analyses"
author: "Maarten Slagter"
output: html_document
---

```{r, include = F, cache = F, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = T, cache.lazy = F, 
                      cache.comments = F, autodep = T, warning = FALSE, 
                      echo = F, error = FALSE)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
source('R/init.R')
```

The relatively small patient numbers per induction arm render the TONIC study
susceptible to potentially confounding variables not being equal between
treatment arms. In an initial clinical response analysis, I already observed
some clinical variables like WHO PS and CA15.3 are unevenly distributed between
treatment arms and correcting for them changed the relative order given to
different induction arms. In the document, I investigate whether we should
similarly control for baseline covariates in NanoString signature fold change
analysis. Do potential confounders increase the likelihood of showing a large
fold change?

```{r, include = F, init, eval = F, warning=FALSE, echo=FALSE}
## Read data
#{
patient_labels <- read.csv(file.path(p_root, 'data-raw',
                                     '180212_clinical_data_all_patients.csv'),
                           dec = ',', sep = ';') %>% as.data.table %>%
  normalize_colnames()
setnames(patient_labels, gsub('\\.', '_', colnames(patient_labels)))
maartenutils::set_dt_types(patient_labels,
                           c('mean_log2_hk' = 'numeric',
                             'tis_score' = 'numeric'))
patient_labels[, patient := paste0('pat_', study_id)]
patient_labels[is.na(patient)]
patient_labels[, table(patient)]
patient_labels <- patient_labels[1:67]
names(patient_labels)
base::grep(x = names(patient_labels), pattern = 'til')
patient_labels[, 'performance_status' := droplevels(performance_status)]
patient_labels[, table(performance_status)]

patient_labels[, .(patient, induction_therapy, performance_status,
                   response_rate, pd_l1_immuuninfiltrate)]
colnames(patient_labels)
# patient_labels[patient == 'pat_31']
## Exclude patient 26, the MMR patient
patient_labels <- patient_labels[-which.max(time_progression_random__weeks)]
# patient_labels[which.max(time_progression_random__weeks)]

# clear_object('brm_tonic', sr)
# clear_all_objects(sr = sr)
```

Looking at the TIS signature only, we see that none of the 4 potential
confounders is strongly or recurrently (across multiple induction arms)
associated with log FC.

```{r, cache = F, warning=FALSE, echo=FALSE}
test_confounding <- function(x_var = 'Baseline',
                             x_label = 'Baseline expression level',
                             tp1 = 'Baseline', tp2 = 'Post-induction',
                             feature = 'tis') {
  c_dat_exp <- prepare_nano_bivariates(x_var = tp1,
                                       y_var = tp2,
                                       colour_var = 'arm') %>%
    merge(clinical_annotation, by = 'patient')

  ggplot(c_dat_exp[variable == feature], aes_string(x = x_var, y = "logfc")) +
    geom_point() +
    facet_grid(~arm) +
    geom_smooth(se = F, span = 2, alpha = .2) +
    xlab(x_label) +
    ylab(sprintf('log FC\n%s vs. %s', tp1, tp2)) +
    ggtitle(feature)
}

plot_confounding_panel <- function(tp1 = 'Baseline', tp2 = 'Post-induction',
                                   feature = 'tis') {
  cowplot::plot_grid(
    test_confounding(x_var = sprintf('`%s`', tp1),
                     x_label = sprintf('%s expression level', tp1),
                     tp1 = tp1, tp2 = tp2, feature = feature),
    test_confounding(x_var = 'ca15_3', x_label = 'Baseline CA15.3',
                     tp1 = tp1, tp2 = tp2, feature = feature),
    test_confounding(x_var = 's_til', x_label = 'Baseline stromal TIL [%]',
                     tp1 = tp1, tp2 = tp2, feature = feature),
    test_confounding(x_var = 'pd_l1_immunoinfiltrate',
                     x_label = 'Baseline PD-L1 [%]',
                     tp1 = tp1, tp2 = tp2, feature = feature),
    ncol = 1)
}
```

Baseline vs. post-induction

```{r, fig.height = 10, warning=FALSE, echo=FALSE}
plot_confounding_panel(tp1 = 'Baseline', tp2 = 'Post-induction',
                       feature = 'tis')
```

Baseline vs. on nivo

```{r, fig.height = 10, warning=FALSE, echo=FALSE}
plot_confounding_panel(tp1 = 'Baseline', tp2 = 'On nivo',
                       feature = 'tis')
```

Post-induction vs. on nivo

```{r, fig.height = 10, warning=FALSE, echo=FALSE}
plot_confounding_panel(tp1 = 'Post-induction', tp2 = 'On nivo',
                       feature = 'tis')
```

It's a bit laborious to be inspecting these plots for all signatures, so I
decided to plot correlation coefficients in heatmaps instead.

```{r, test_confounding_heatmap, cache = F, warning=FALSE, echo=FALSE}
test_confounding_heatmap <- function(tp1 = 'Baseline', tp2 = 'Post-induction') {
  c_dat_exp <- prepare_nano_bivariates(x_var = tp1,
                                       y_var = tp2,
                                       colour_var = 'arm') %>%
    merge(clinical_annotation, by = 'patient')
  # x_label = sprintf('%s expression level', tp1)
  # x_label = 'Baseline PD-L1 [%]'
  # comp_vars <- c(sprintf('`%s`', tp1), 'ca15_3', 's_til',
  #                'pd_l1_immunoinfiltrate')
  comp_vars <- c(tp1, 'ca15_3', 's_til', 'pd_l1_immunoinfiltrate')

  plots <- lapply(c_dat_exp[, unique(variable)], function(feat) {
    p_dat <- rbindlist(lapply(auto_name(comp_vars), function(comp_var) {
        c_dat_exp[variable == feat,
                  .('cor' = cor(logfc, get(comp_var), method = 'spearman'),
                    'var' = comp_var,
                    'feat' = feat),
                  by = arm]
        }), fill = T)

    ggplot(p_dat, aes(x = arm, y = var, fill = cor,
                      label = sprintf('%.2f', cor))) +
      geom_raster() +
      geom_text(size = 2) +
      scale_fill_gradient2(name = 'Spearman correlation',
                           low = scales::muted('red'),
                           high = scales::muted('blue'),
                           limits = c(-1, 1),
                           na.value = 'white') +
      ggtitle(feat) +
      rotate_x_labels(45) +
      xlab('') + ylab('')
  })

  plot_panel_layout(plots, ncol = 2, nrow = 3,
                    labels = NULL,
                    plot_direct = F,
                    # filename = file.path(plot_dir, 'baseline_logfc_cors.pdf'),
                    clear_redundant_labels = T,
                    clear_redundant_legends = T)
}

```

Baseline vs. post-induction

```{r, fig.height = 12, fig.width = 12, warning=FALSE, echo=FALSE, dependson='test_confounding_heatmap'}
devtools::load_all(file.path('~/libs', 'maartenutils'))
p <- test_confounding_heatmap(tp1 = 'Baseline', tp2 = 'Post-induction')
```

Baseline vs. on nivo

```{r, fig.height = 12, fig.width = 12, warning=FALSE, echo=FALSE, dependson='test_confounding_heatmap'}
p <- test_confounding_heatmap(tp1 = 'Baseline', tp2 = 'On nivo')
```

Post-induction vs. on nivo

```{r, fig.height = 12, fig.width = 12, warning=FALSE, echo=FALSE, dependson='test_confounding_heatmap'}
p <- test_confounding_heatmap(tp1 = 'Post-induction', tp2 = 'On nivo')
```

The previous is still a bit daunting to look at. I wondered whether some
baseline confounders are more likely to be associated with on another, across
gene signatures.  As shown below, a gene signature's baseline levels
anti-correlate with the FC that's achieved by all induction therapies. This can
probably be understood by the log transformation that is applied to the gene
signature scores. This results in increasing difficulty for a signature score to
increase with high baseline levels. Shown below are distributions of correlation
coefficients for all of the gene signature fold changes with the four indicated
variables (in panel rows).

```{r, cache = F, warning=FALSE, echo=FALSE}
test_confounding_cor_hists <- function(tp1 = 'Baseline', 
                                       tp2 = 'Post-induction') {
  c_dat_exp <- prepare_nano_bivariates(x_var = tp1,
                                       y_var = tp2,
                                       colour_var = 'arm') %>%
    merge(clinical_annotation, by = 'patient')

  comp_vars <- c(tp1, 'ca15_3', 's_til', 'pd_l1_immunoinfiltrate')

  t_dat <- rbindlist(lapply(c_dat_exp[, unique(variable)], function(feat) {
    rbindlist(lapply(auto_name(comp_vars), function(comp_var) {
      c_dat_exp[variable == feat,
                .('cor' = cor(logfc, get(comp_var), method = 'spearman'),
                  'var' = comp_var,
                  'feat' = feat),
                by = arm]
      }), fill = T) 
  }), fill = T)

  t_dat[, .('frac_pos' = .SD[cor > 0, .N] / .SD[!is.na(cor), .N]), 
        by = c('var', 'arm')]
  
  ggplot(t_dat, aes(x = cor)) + 
    facet_grid(var ~ arm, labeller = label_value) +
    geom_vline(xintercept = 0, colour = 'grey85', type = 'dashed') +
    geom_histogram() + 
    xlab('Spearman correlation coefficients') +
    ylab('Number of sigs') +
    geom_vline(data = t_dat[, .('median' = median(cor, na.rm = T)), 
                            by = c('var', 'arm')],
               aes(xintercept = median), colour = 'red')

}
```

Baseline vs. post-induction

```{r, fig.height = 14, fig.width = 17, warning=FALSE, echo=FALSE, dependson='test_confounding_heatmap'}
test_confounding_cor_hists(tp1 = 'Baseline', tp2 = 'Post-induction')
```

Baseline vs. on nivo

```{r, fig.height = 14, fig.width = 17, warning=FALSE, echo=FALSE, dependson='test_confounding_heatmap'}
test_confounding_cor_hists(tp1 = 'Baseline', tp2 = 'On nivo')
```

Post-induction vs. on nivo

```{r, fig.height = 14, fig.width = 17, warning=FALSE, echo=FALSE, dependson='test_confounding_heatmap'}
test_confounding_cor_hists(tp1 = 'Post-induction', tp2 = 'On nivo')
```

## Enrichment of signatures amongst those for which correlations exist?

```{r, cache = F, warning=FALSE, echo=FALSE}
## TODO FINISH ME
check_confounding_cor_hists <- function(tp1 = 'Baseline', 
                                       tp2 = 'Post-induction') {
  c_dat_exp <- prepare_nano_bivariates(x_var = tp1,
                                       y_var = tp2,
                                       colour_var = 'arm') %>%
    merge(clinical_annotation, by = 'patient')

  comp_vars <- c(tp1, 'ca15_3', 's_til', 'pd_l1_immunoinfiltrate')

  t_dat <- rbindlist(lapply(c_dat_exp[, unique(variable)], function(feat) {
    rbindlist(lapply(auto_name(comp_vars), function(comp_var) {
      c_dat_exp[variable == feat,
                .('cor' = cor(logfc, get(comp_var), method = 'spearman'),
                  'var' = comp_var,
                  'feat' = feat),
                by = arm]
      }), fill = T) 
  }), fill = T)

  ggplot(t_dat, aes(x = cor)) + 
    facet_grid(var ~ arm, labeller = label_value) +
    geom_vline(xintercept = 0, colour = 'grey85', type = 'dashed') +
    geom_histogram() + 
    xlab('Spearman correlation coefficients') +
    ylab('Number of sigs') +
    geom_vline(data = t_dat[, .('median' = median(cor, na.rm = T)), 
                            by = c('var', 'arm')],
               aes(xintercept = median), colour = 'red')

}
check_confounding_cor_hists()
```
